---
title: "Catalog_inflammation"
author: "Moises A. Rojas"
date: "08/07/2026"
output:
  html_document:
    toc: true
---

```{r setup}
suppressMessages(library(tidyverse)) #tidy ecosystem
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(ggplot2)) #plotting
suppressMessages(library(patchwork)) #side-by-side plotting
suppressMessages(library(RColorBrewer)) #color palettes
suppressMessages(library(pals)) #more colors
suppressWarnings(library(colorspace)) #color gradients
suppressWarnings(library(colorRamp2)) #color gradients
suppressMessages(library(biomaRt)) #ensembl convertion
suppressMessages(library(clusterProfiler)) #KEGG enrichment
suppressMessages(library(enrichplot)) #enrichment plots
suppressMessages(library(KEGGREST)) #KEGG id conversion
suppressMessages(library(KEGGgraph)) #KEGG network
suppressMessages(library(Rgraphviz)) #graph viz
suppressMessages(library(gtexr)) #GTEx portal API
suppressPackageStartupMessages(library(ComplexHeatmap)) #heatmap plot
suppressMessages(library(mclust)) #Gaussian modeling

# ggplot(...) + ... + theme_local()
theme_local <- function() {
  theme_classic() +
    theme(
      legend.box.spacing = unit(0.5, "pt"), #space between plot and legend
      legend.key.spacing.x = unit(0.1, "pt"), #space between columns
      legend.key.spacing.y = unit(0.1, "pt"), #space between rows
      legend.spacing.y = unit(0.75, "pt"), #space between legend titles
      legend.key.height = unit(3, units = "mm"),
      legend.key.width = unit(4, units = "mm")
    )
}
```

## A set of inflammatory genes

Convert the initial set of genes using the HGNC symbols (n = 85).

```{r convert}
# Connect to the human Ensembl database
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Read file of selected genes as input
genes_file <- read.csv("inflammation_genes.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Create a list of genes
genes_list <- genes_file %>% pull(HGNC_symbol)

# Map query genes to get IDs from Uniprot and GeneCards
genes_query <- getBM(attributes = c("hgnc_symbol", "uniprotswissprot", "genecards"), #reactome, go_id..
                     filter = "hgnc_symbol", values = genes_list, mart = ensembl)
##---ensure full conversion here
```

We check if some genes are missing after convertion for manual curation, then save results to file.

```{r check}
# Filter empty rows and print results
genes_query <- genes_query %>% filter(uniprotswissprot != "")
print(genes_query)

print(genes_file[which(!(genes_file$HGNC_symbol %in% genes_query$hgnc_symbol)), ])
genes_query <- full_join(genes_file, genes_query, by = join_by(HGNC_symbol == hgnc_symbol))
#write.table(genes_query, "output.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

This step will allow us to update our internal dataset by updating Sheet nº2 in the following file: https://docs.google.com/spreadsheets/d/1DkwHyp2xs9bxtTHlldugihmQJX4yX36Zi3WXIjvqDoM/edit?usp=sharing.

## KEGG enrichment analysis

The curated_genes.tsv file contains a preprocessed list of inflammatory genes annotated with UniProt and GeneCards identifiers. The UniProt IDs are then used to retrieve the corresponding KEGG IDs, useful for enrichment analyses.

```{r keggread}
# Read the curated list of genes
genes_file <- read.csv("enrichKEGG/data/curated_genes.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
genes_uniprot <- genes_file %>% pull(uniprot)
print(length(genes_uniprot))

# Print database version -> Last update 2026/07/13
cat(readLines("https://rest.kegg.jp/info/pathway/")[[3]])
```

### clusterProfiler

We perform a Over-Representation Analysis (ORA) using the Uniprot IDs.

```{r keggclus}
# Run enrichment
kegg_res <- enrichKEGG(gene = genes_uniprot, organism = "hsa", keyType = "uniprot",
                       pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 0.05)
```

Create some plots to explore the enrichment.

```{r plotclus}
# Plot enriched dotplot
#use: label_format = funlab
#x = ~ -log(p.adjust),
#check in output: only pathways or phenotypes?
dotplot(kegg_res, showCategory = 20, font.size = 10)

# Plot barplot
kegg_res %>%
  mutate(qscore = -log(p.adjust, base = 10)) %>%
  barplot(x = "qscore", showCategory = 20, font.size = 10)

# Plot the most significant terms and their relationships
#kegg_simmat <- pairwise_termsim(kegg_res, method = "JC")
#ssplot(kegg_simmat)
```

### Local enrichment

First, we convert list of genes from Uniprot IDs to KEGG IDs.

```{r uni2keggid}
kegg_genes <- bitr_kegg(geneID = genes_uniprot, fromType = "uniprot",
                        toType = "kegg", organism = "hsa")

# Check duplicated entries (genes associated with more than one KEGG ID)
print(unique(kegg_genes$uniprot[duplicated(kegg_genes$uniprot)]))
#kegg_genes <- kegg_genes %>% distinct(kegg, .keep_all = TRUE)

# Update the initial list with KEGG IDs
genes_file <- left_join(genes_file, kegg_genes, by = "uniprot")
```

Our list now contains a total of 87 genes because two of them (P0C0L4 and P0C0L5) are associated with two KEGG IDs each.

#### KEGG KO approach

Next, we use those IDs to get the KEGG KOs.

```{r id2ko}
# Append organism prefix to each entry
genes_map <- kegg_genes %>% pull(kegg)
genes_map <- append(genes_map, list(paste0("hsa:", genes_map)))
genes_map <- unlist(tail(genes_map, 1)) #keep last position
head(genes_map)

# Convert KEGG IDs to KOs -> https://www.genome.jp/dbget-bin/www_bget?hsa:721
kegg_kos <- keggLink("ko", genes_map)

# Check duplicated entries
print(kegg_kos[duplicated(kegg_kos)])

# Create table and export it to file
kegg_id_ko <- tibble(ID = names(kegg_kos), KO = unname(kegg_kos))
kegg_id_ko <- kegg_id_ko %>%
  dplyr::mutate(ID = str_replace(ID, "hsa:", ""), #remove prefix
                KO = str_replace(KO, "ko:", ""))
#write.table(kegg_id_ko, "enrichKEGG/data/KEGG_id2ko.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

# Update the initial list with KEGG KOs
genes_file <- genes_file %>%
  dplyr::rename(kegg_id = kegg) %>%
  left_join(kegg_id_ko %>% distinct(ID, KO), by = join_by(kegg_id == ID)) %>%
  dplyr::rename(kegg_ko = KO)
```

Among these, the same KO (K03989) is assigned to two different KEGG IDs, both corresponding to complement component 4 (C4; C4A and C4B). So, we have 84 unique KOs.
The next step is to map the KEGG pathways based on the converted KOs to determine which pathways are represented by our set of genes.

```{r mapkos}
kegg_ko_path <- keggLink("pathway", kegg_kos)
kegg_data <- tibble(KO = names(kegg_ko_path), pathway = unname(kegg_ko_path))

# Which KOs are missing? (not associated with any pathway)
print(kegg_kos[!kegg_kos %in% kegg_data$KO])
print(length(unique(kegg_data$KO))) #81
```

We observe that three KOs (K04245, K11994, and K23267) are not associated with any KEGG pathway, and were therefore excluded from the pathway-based analysis. This reduces the set to 81 KOs.

```{r ko2path}
# Modify and filter columns
kegg_data <- kegg_data %>%
  dplyr::mutate(KO = str_replace(KO, "ko:", "")) %>%
  dplyr::mutate(pathway = str_replace(pathway, "path:map", "")) %>% #keep map?
  dplyr::filter(!grepl("ko", pathway))

# Check the pathway names from the KEGG database (v.042022)
#pathway_list <- read.csv("enrichKEGG/data/pathway.list.mod.tsv", sep = "\t", header = TRUE,
#                         colClasses = c("character", "character"), stringsAsFactors = FALSE)
#print(kegg_categories[!(kegg_categories$id %in% pathway_list$id), ]) #ie. 04148 --> Efferocytosis

# Get KEGG pathway names from API
kegg_categories <- clusterProfiler:::kegg_category_data()
kegg_data <- left_join(kegg_data, kegg_categories, by = join_by(pathway == id))

# Process NAs manually
#https://www.kegg.jp/pathway/map<pathway>
#https://www.genome.jp/kegg/docs/upd_kegg.html
print(kegg_data %>% filter(is.na(name)) %>% distinct(pathway))
kegg_data <- kegg_data %>%
  dplyr::mutate(name = str_replace(name, "Efferocytosis New!", "Efferocytosis")) %>% #edit name
  dplyr::mutate(name = case_when(pathway == "04081" ~ "Hormone signaling",
                                 pathway == "04082" ~ "Neuroactive ligand signaling",
                                 pathway == "04517" ~ "IgSF CAM signaling",
                                 pathway == "04518" ~ "Integrin signaling",
                                 pathway == "04519" ~ "Cadherin signaling",
                                 TRUE ~ name)) %>%
  dplyr::mutate(subcategory = case_when(pathway == "04081" ~ "Signaling molecules and interaction",
                                        pathway == "04082" ~ "Signaling molecules and interaction",
                                        pathway == "04517" ~ "Signaling molecules and interaction",
                                        pathway == "04518" ~ "Signaling molecules and interaction",
                                        pathway == "04519" ~ "Signaling molecules and interaction",
                                        TRUE ~ subcategory)) %>%
  dplyr::mutate(category = case_when(pathway == "04081" ~ "Environmental Information Processing",
                                     pathway == "04082" ~ "Environmental Information Processing",
                                     pathway == "04517" ~ "Environmental Information Processing",
                                     pathway == "04518" ~ "Environmental Information Processing",
                                     pathway == "04519" ~ "Environmental Information Processing",
                                     TRUE ~ category))

#get KO name? *** ej. K03989 -> C4; complement component 4
```

We plot the global distribution of KEGG pathways.

```{r plotdistko}
kegg_data %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(n_path = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n_path)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n_path, y = fct_reorder(name, n_path))) + #, fill = name
  geom_col(position = "dodge2", fill = "#459682") +
  #scale_fill_discrete_diverging(name = "", palette = "Purple-Green") +
  geom_text(aes(label = n_path), vjust = 0.5, hjust = -0.25, size = 3) +
  labs(title = "KEGG pathways distribution (top 30)", x = "# of genes", y = "Pathway name") +
  theme_classic() +
  theme(legend.position = "none")

# Set colors for subcategories
colors_subcat <- brewer.dark2(length(unique(kegg_data$subcategory)))

# Group by subcategory
kegg_data %>%
  dplyr::group_by(pathway, subcategory, name) %>%
  dplyr::summarise(n_path = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n_path)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n_path, y = fct_reorder(name, n_path), fill = subcategory)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = n_path), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Subcategory", values = colors_subcat) +
  labs(title = "KEGG pathways per subcategory (top 30)", x = "# of genes", y = "Pathway name") +
  theme_local()

# Set colors for categories
colors_cat <- colorRampPalette(brewer.pal(8, "Set2"))(length(unique(kegg_data$category)))

# Group by category
kegg_data %>%
  dplyr::group_by(pathway, category, name) %>%
  dplyr::summarise(n_path = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n_path)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n_path, y = fct_reorder(name, n_path), fill = category)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = n_path), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Category", values = colors_cat) +
  labs(title = "KEGG pathways per category (top 30)", x = "# of genes", y = "Pathway name") +
  theme_local()
```

At this point, the dataset comprises a total of 215 pathways. But, there are some pathways derived from ortholog genes belonging to other organisms. For instance, "MAPK signaling pathway - fly" (04013), "Meiosis - yeast" (04113), "Longevity regulating pathway - worm" (04212), or "Apoptosis - fly" (04214).

#### KEGG ID approach

Lets try with the mapping of KEGG pathways by using directly the IDs (skipping ortholog (KO) convertion)...

```{r mapids}
kegg_id2path <- keggLink("pathway", genes_map)
kegg_data_id <- tibble(ID = names(kegg_id2path), pathway = unname(kegg_id2path))

# Which KEGG IDs are missing? (not associated with any pathway)
print(genes_map[!genes_map %in% kegg_data_id$ID])
print(length(unique(kegg_data_id$ID))) #83
```

We observe the same three genes (ID 1240: K04245, ID 1773: K11994, and ID 325: K23267) that are not associated with any KEGG pathway. This reduces the set to 83 KEGG IDs.

```{r id2path}
# Modify and filter columns
kegg_data_id <- kegg_data_id %>%
  dplyr::mutate(ID = str_replace(ID, "hsa:", "")) %>%
  dplyr::mutate(pathway = str_replace(pathway, "path:hsa", ""))

# Get KEGG pathway names from API
kegg_data_id <- left_join(kegg_data_id, kegg_categories, by = join_by(pathway == id))

# Curate NAs manually
#https://www.kegg.jp/pathway/map<pathway>
#https://www.genome.jp/kegg/docs/upd_kegg.html
print(kegg_data_id %>% dplyr::filter(is.na(name)) %>% distinct(pathway))
kegg_data_id <- kegg_data_id %>%
  dplyr::mutate(name = str_replace(name, "Efferocytosis New!", "Efferocytosis")) %>% #edit name
  dplyr::mutate(name = case_when(pathway == "04081" ~ "Hormone signaling",
                                 pathway == "04082" ~ "Neuroactive ligand signaling",
                                 pathway == "04517" ~ "IgSF CAM signaling",
                                 pathway == "04518" ~ "Integrin signaling",
                                 pathway == "04519" ~ "Cadherin signaling",
                                 TRUE ~ name)) %>%
  dplyr::mutate(subcategory = case_when(pathway == "04081" ~ "Signaling molecules and interaction",
                                        pathway == "04082" ~ "Signaling molecules and interaction",
                                        pathway == "04517" ~ "Signaling molecules and interaction",
                                        pathway == "04518" ~ "Signaling molecules and interaction",
                                        pathway == "04519" ~ "Signaling molecules and interaction",
                                        TRUE ~ subcategory)) %>%
  dplyr::mutate(category = case_when(pathway == "04081" ~ "Environmental Information Processing",
                                     pathway == "04082" ~ "Environmental Information Processing",
                                     pathway == "04517" ~ "Environmental Information Processing",
                                     pathway == "04518" ~ "Environmental Information Processing",
                                     pathway == "04519" ~ "Environmental Information Processing",
                                     TRUE ~ category))

print(length(unique(kegg_data_id$pathway))) #208
print(length(unique(kegg_data_id$category))) #6
print(length(unique(kegg_data_id$subcategory))) #32
```

Considering this approach, the dataset comprises now a total of 208 pathways, distributed across 6 categories and 32 subcategories according to KEGG classification.

```{r plotdistid}
#, fig.width = 10, fig.height = 8, out.width = "100%"
kegg_data_id %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(n = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n, y = fct_reorder(name, n))) + #, fill = name
  geom_col(position = "dodge2", fill = "#459682") +
  geom_text(aes(label = n), vjust = 0.5, hjust = -0.25, size = 3) +
  labs(title = "KEGG pathways distribution (top 30)", x = "# of genes", y = "Pathway name") +
  theme_classic() +
  theme(legend.position = "none")

# Set colors for subcategories
colors_subcat <- brewer.dark2(length(unique(kegg_data_id$subcategory)))

# Group by subcategory
p1 <- kegg_data_id %>%
  dplyr::group_by(pathway, subcategory, name) %>%
  dplyr::summarise(n = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n, y = fct_reorder(name, n), fill = subcategory)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = n), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Subcategory", values = colors_subcat) +
  labs(title = "KEGG pathways per subcategory (top 30)", x = "# of genes", y = "Pathway name") +
  theme_local()

# Set colors for categories
colors_cat <- colorRampPalette(brewer.pal(8, "Set2"))(length(unique(kegg_data_id$category)))

# Group by category
p2 <- kegg_data_id %>%
  dplyr::group_by(pathway, category, name) %>%
  dplyr::summarise(n = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n)) %>%
  dplyr::slice(1:30) %>% #top 30
  ggplot(aes(x = n, y = fct_reorder(name, n), fill = category)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = n), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Category", values = colors_cat) +
  labs(title = "KEGG pathways per category (top 30)", x = "# of genes", y = "Pathway name") +
  theme_local()

# Agglom. plots
(p1 / p2) + plot_annotation(tag_levels = list(c("a", "b")))
ggsave("enrichKEGG/plots/plot_KEGG_pathways.png", height = 11, width = 9)
```

The plots above exhibit the global distribution of the 30 most represented KEGG pathways for the input dataset of inflammatory genes.

#### Enrichment

We calculate the enrichment-like ratio of genes per pathway.

```{r enrich}
kegg_enrichment <- kegg_data_id %>%
  dplyr::group_by(ID, pathway, name) %>% #separate by IDs
  dplyr::summarise(n_paths = n(), .groups = "drop") %>%
  ungroup() %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(across(ID, \(x) paste(na.omit(x), collapse = ",")),
                   count = n(), .groups = "drop") %>% #collapse IDs by pathway and count them
  ungroup() %>%
  dplyr::rename(geneIDs = ID) %>%
  dplyr::mutate(ratio = count / length(unique(kegg_data_id$ID))) #gene ratio is count/total genes
```

Download the complete KEGG human pathway-to-gene mapping.

```{r kegguniv}
kegg_all_hsa <- keggLink("pathway", "hsa")
kegg_mapping <- data.frame(gene_id = gsub("", "", names(kegg_all_hsa)), #hsa:
                           path_id = gsub("path:hsa", "", unname(kegg_all_hsa)))
print(length(unique(kegg_mapping$gene_id))) #total genes: 9421
rm(kegg_all_hsa) #remove

# In case of updating the vector:
#setdiff(names(kegg_all_hsa), kegg_mapping$gene_id)
```

We define the actual background universe of pathways and genes per pathway.

```{r bgpath}
# Define the total number of all unique genes in the background distribution
bg_genes <- length(unique(kegg_mapping$gene_id)) #N
kegg_map_path <- kegg_mapping %>%
  dplyr::group_by(path_id) %>%
  dplyr::summarise(M = n()) #number of genes in the background universe (per pathway)

# Merge background data with the enrichment
kegg_enrichment <- kegg_enrichment %>%
  left_join(kegg_map_path, by = join_by(pathway == path_id)) %>%
  dplyr::mutate(BgRatio = paste0(M, "/", bg_genes)) #calculate the BgRatio = M/N
```

Using the background universe, a hypergeometric distribution is applied to calculate the p-value. This statistical test evaluates the significance of the overlap between the gene set and the universe.

```{r bgratio}
# Define the number of genes in the gene set
n_paths <- length(unique(kegg_data_id$pathway)) #n
#phyper(k-1, M, N-M, n, lower.tail = FALSE)
#use correction methods: such as Bonferroni or false discovery rate (FDR)
kegg_enrichment <- kegg_enrichment %>%
  dplyr::mutate(pvalue = phyper(count - 1, M, bg_genes - M, n_paths, lower.tail = FALSE)) %>%
  dplyr::mutate(p.adjust = p.adjust(pvalue, method = "BH"))

# Get and sort the significant pathways
print(kegg_enrichment %>%
        dplyr::filter(pvalue <= 0.05) %>%
        dplyr::select(name, ratio, p.adjust) %>%
        dplyr::arrange(desc(ratio)))
```

We create a dotplot of the local enrichment.

```{r plotenrich}
# Mark significant pathways
kegg_enrichment <- kegg_enrichment %>%
  left_join(kegg_paths %>% dplyr::select(pathway, category, subcategory), by = "pathway") %>%
  dplyr::mutate(signif = ifelse(p.adjust < 0.05, "*", ""))

# Plot top 40 pathways
kegg_enrichment %>%
  dplyr::arrange(desc(ratio)) %>%
  dplyr::slice(1:40) %>% #top 40
  ggplot(aes(x = ratio, y = fct_reorder(name, ratio))) +
  geom_point(aes(size = count, color = p.adjust), alpha = 0.85) +
  geom_text(aes(label = signif, x = -0.005), size = 3.5, color = "#3c4d76") + #p.adj
  scale_y_discrete(expand = c(0.05, 0)) + #add y-axis space
  scale_color_gradient(name = "p.adjust", low = "red", high = "blue") +
  guides(size = guide_legend(title = "# genes", order = 1, reverse = TRUE,
                             override.aes = list(color = "black", alpha = 0.85))) +
  labs(title = "KEGG local enrichment (top 40)",
       subtitle = paste0("Total pathways: ", nrow(kegg_enrichment)),
       x = "Gene ratio", y = "Pathway name") +
  #facet_wrap(~category, nrow = 2, axes = "all_x", scales = "free_y") +
  theme_local()
ggsave("enrichKEGG/plots/plot_KEGG_enrichment.png", height = 6, width = 7)
#ggsave("enrichKEGG/plots/plot_KEGG_enrichment.pdf", height = 6, width = 7)

# Export dataset
#write.table(kegg_enrichment, "enrichKEGG/output_data/KEGG_enrichment.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

In the previous plot of the 40 most enriched pathways, almost all pathways showed statistically significant p-values. Among the top three pathways, the "Pathways in cancer" has a high enrichment ratio despite being less represented in the background universe (p = 0.147). In total, there are 53/208 significant pathways.

```{r enrichsignif}
# Plot significant pathways only
kegg_enrichment %>%
  dplyr::arrange(desc(ratio)) %>%
  dplyr::filter(signif == "*") %>%
  ggplot(aes(x = ratio, y = fct_reorder(name, ratio))) +
  geom_point(aes(size = count, color = p.adjust), alpha = 0.85) +
  scale_y_discrete(expand = c(0.05, 0)) + #add y-axis space
  scale_color_gradient(name = "p.adjust", low = "red", high = "blue") +
  guides(size = guide_legend(title = "# genes", order = 1, reverse = TRUE,
                             override.aes = list(color = "black", alpha = 0.85))) +
  labs(title = "KEGG local enrichment",
       subtitle = paste0("Significant enriched pathways: ",
                         kegg_enrichment %>% dplyr::filter(signif == "*") %>% nrow()),
       x = "Gene ratio", y = "Pathway name") +
  theme_local()
ggsave("enrichKEGG/plots/KEGG_enrichment_signif.pdf", height = 8, width = 7)
```

List all genes per pathways.

```{r enrichgenes}
kegg_engenes <- kegg_enrichment %>%
  separate_longer_delim(cols = geneIDs, delim = ",") %>% #split 114548,1535,1536,29108,..., to rows
  group_by(geneIDs) %>%
  summarise(genes = n()) %>%
  left_join(kegg_data_id, by = join_by(geneIDs == ID)) #add KEGG data

# Plot all genes across pathways
kegg_engenes %>%
  distinct(geneIDs, .keep_all = TRUE) %>% #collapse list
  left_join(genes_file, by = join_by(geneIDs == kegg_id)) %>% #add gene data from the input list
  dplyr::mutate(gene_lab = paste0(HGNC_symbol, " (KEGG ID: ", geneIDs, ") - ", Main_function)) %>% #label
  arrange(desc(genes)) %>%
  ggplot(aes(x = genes, y = fct_reorder(gene_lab, genes))) +
  geom_point(aes(size = genes, color = category), alpha = 0.75) + #"#e31a1c"
  geom_text(aes(label = genes), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_y_discrete(expand = c(0.025, 0)) + #add y-axis space
  scale_fill_manual(name = "Category", values = colors_cat) +
  guides(size = "none") +
  labs(title = "Inflammatory genes", x = "# of pathways", y = "Gene name and function") +
  theme_local()
ggsave("enrichKEGG/plots/plot_KEGG_key_genes.png", height = 15, width = 11)
```

The plot above depicts the global distribution of the 83 inflammatory genes, summarizing the pathways identified by the enrichment analysis.

## KEGG statistics

We explore some data in the KEGG database: the average number of genes per pathways.

```{r keggstats}
# Count genes per pathway
kegg_genes_per_pathway <- kegg_mapping %>%
  dplyr::group_by(path_id) %>%
  dplyr::summarise(n_genes = n_distinct(gene_id))

# Plot histogram
kegg_genes_per_pathway %>%
  ggplot(aes(x = n_genes)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue", alpha = 0.75) + #binwidth = 0.2
  stat_bin(aes(label = after_stat(count)), bins = 30, geom = "text", size = 3, vjust = -0.5) +
  scale_y_continuous(limits = c(0, 150)) +
  labs(title = "Histogram of genes across KEGG pathways",
       subtitle = paste0("Total pathways: ", length(unique(kegg_mapping$path_id)),
                         "\nAverage number of genes per pathway: ",
                         round(mean(kegg_genes_per_pathway$n_genes), 2)),
       x = "Number of genes per pathway", y = "Number of pathways") +
  theme_classic()
ggsave("enrichKEGG/plots/plot_KEGG_stats.png", height = 5, width = 7)

# Plot histogram without outlier
kegg_genes_per_pathway %>%
  dplyr::filter(n_genes < 1000) %>%
  ggplot(aes(x = n_genes)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue", alpha = 0.75) +
  stat_bin(aes(label = after_stat(count)), bins = 30, geom = "text", size = 3, vjust = -0.5) +
  geom_vline(aes(xintercept = median(kegg_genes_per_pathway$n_genes)),
             linetype = "dashed", linewidth = 0.2) +
  annotate("text", x = median(kegg_genes_per_pathway$n_genes), y = 60, color = "black",
           label = paste0("Median: ", round(median(kegg_genes_per_pathway$n_genes), 2)),
           size = 4, hjust = 0) +
  scale_x_continuous(limits = c(0, 600)) +
  scale_y_continuous(limits = c(0, 60)) +
  labs(x = "Number of genes per pathway", y = "Number of pathways") +
  theme_classic()
ggsave("enrichKEGG/plots/plot_KEGG_stats_v2.png", height = 5, width = 7)

# Download the updated list of KEGG pathways
#kegg_paths <- names(keggList("pathway", "hsa"))
#kegg_paths <- tibble(pathway = kegg_paths)
#kegg_paths <- kegg_paths %>%
#  dplyr::mutate(pathway = str_replace(pathway, "hsa", "")) %>% #remove prefix
#  left_join(kegg_categories, by = join_by(pathway == id))
#kegg_classes <- keggGet("br:br08901")
kegg_paths <- read.csv("https://rest.kegg.jp/list/pathway/hsa", sep = "\t", header = FALSE)
glimpse(kegg_paths)
kegg_paths <- kegg_paths %>%
  dplyr::rename(pathway = V1, name = V2) %>%
  dplyr::mutate(pathway = str_replace(pathway, "hsa", "")) %>% #remove prefix
  dplyr::mutate(name = str_replace_all(name, fixed(" - Homo sapiens (human)"), "")) #remove suffix

# Add categories and subcategories
kegg_paths <- kegg_paths %>%
  left_join(kegg_categories %>% dplyr::select(id, category, subcategory),
            by = join_by(pathway == id))
print(kegg_paths %>% dplyr::filter(is.na(category)))

# Manually update classes
kegg_paths <- kegg_paths %>%
  dplyr::mutate(category = case_when(pathway == "01320" ~ "Metabolism",
                                     pathway == "03272" ~ "Genetic Information Processing",
                                     pathway == "03273" ~ "Genetic Information Processing",
                                     pathway == "04081" ~ "Environmental Information Processing",
                                     pathway == "04082" ~ "Environmental Information Processing",
                                     pathway == "04517" ~ "Environmental Information Processing",
                                     pathway == "04518" ~ "Environmental Information Processing",
                                     pathway == "04519" ~ "Environmental Information Processing",
                                     pathway == "04156" ~ "Cellular Processes",
                                     pathway == "04820" ~ "Cellular Processes",
                                     pathway == "04980" ~ "Organismal Systems",
                                     pathway == "04981" ~ "Organismal Systems",
                                     pathway == "04382" ~ "Organismal Systems",
                                     TRUE ~ category)) %>%
  dplyr::mutate(subcategory = case_when(pathway == "01320" ~ "Energy metabolism",
                                        pathway == "03272" ~ "Information processing in viruses",
                                        pathway == "03273" ~ "Information processing in viruses",
                                        pathway == "04081" ~ "Signaling molecules and interaction",
                                        pathway == "04082" ~ "Signaling molecules and interaction",
                                        pathway == "04517" ~ "Signaling molecules and interaction",
                                        pathway == "04518" ~ "Signaling molecules and interaction",
                                        pathway == "04519" ~ "Signaling molecules and interaction",
                                        pathway == "04156" ~ "Cell growth and death",
                                        pathway == "04820" ~ "Cell motility",
                                        pathway == "04980" ~ "Digestive system",
                                        pathway == "04981" ~ "Digestive system",
                                        pathway == "04382" ~ "Development and regeneration",
                                        TRUE ~ subcategory))

# Export curated pathways to file
#write.table(kegg_paths, "enrichKEGG/output_data/KEGG_pathways_hsa.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

genexpr_all_path <- genexpr_all_path %>%
  dplyr::mutate(name = str_replace(name, "Efferocytosis New!", "Efferocytosis"),
                name = str_replace(name, "Virion - Rotavirus New!", "Virion - Rotavirus"),
                name = str_replace(name, "Virion - Ebolavirus and Lyssavirus New!",
                                   "Virion - Ebolavirus and Lyssavirus")) %>% #edit names
  dplyr::rename(pathway = path_id) %>% #rename
  dplyr::mutate(category = case_when(pathway == "01320" ~ "Metabolism",
                                     pathway == "03272" ~ "Genetic Information Processing",
                                     pathway == "03273" ~ "Genetic Information Processing",
                                     pathway == "04081" ~ "Environmental Information Processing",
                                     pathway == "04082" ~ "Environmental Information Processing",
                                     pathway == "04156" ~ "Cellular Processes",
                                     pathway == "04382" ~ "Organismal Systems",
                                     pathway == "04517" ~ "Environmental Information Processing",
                                     pathway == "04518" ~ "Environmental Information Processing",
                                     pathway == "04519" ~ "Environmental Information Processing",
                                     pathway == "04820" ~ "Cellular Processes",
                                     pathway == "04980" ~ "Organismal Systems",
                                     pathway == "04981" ~ "Organismal Systems",
                                     TRUE ~ category)) %>%
  dplyr::mutate(subcategory = case_when(pathway == "01320" ~ "Energy metabolism",
                                        pathway == "03272" ~ "Information processing in viruses",
                                        pathway == "03273" ~ "Information processing in viruses",
                                        pathway == "04081" ~ "Signaling molecules and interaction",
                                        pathway == "04082" ~ "Signaling molecules and interaction",
                                        pathway == "04156" ~ "Cell growth and death",
                                        pathway == "04382" ~ "Development and regeneration",
                                        pathway == "04517" ~ "Signaling molecules and interaction",
                                        pathway == "04518" ~ "Signaling molecules and interaction",
                                        pathway == "04519" ~ "Signaling molecules and interaction",
                                        pathway == "04820" ~ "Cell motility",
                                        pathway == "04980" ~ "Digestive system",
                                        pathway == "04981" ~ "Digestive system",
                                        TRUE ~ subcategory)) %>%
  dplyr::mutate(name = case_when(pathway == "01320" ~ "Sulfur cycle",
                                 pathway == "03272" ~ "Virion - Hepatitis viruses",
                                 pathway == "03273" ~ "Virion - Lassa virus and SFTS virus",
                                 pathway == "04081" ~ "Hormone signaling",
                                 pathway == "04082" ~ "Neuroactive ligand signaling",
                                 pathway == "04156" ~ "Integrated stress response (ISR) signaling pathway",
                                 pathway == "04382" ~ "Cornified envelope formation",
                                 pathway == "04517" ~ "IgSF CAM signaling",
                                 pathway == "04518" ~ "Integrin signaling",
                                 pathway == "04519" ~ "Cadherin signaling",
                                 pathway == "04820" ~ "Cytoskeleton in muscle cells",
                                 pathway == "04980" ~ "Cobalamin transport and metabolism",
                                 pathway == "04981" ~ "Folate transport and metabolism",
                                 TRUE ~ name))

# Plot stats of categories
colors_keggcat <- colorRampPalette(brewer.pal(8, "Set2"))(length(unique(kegg_paths$category)))
kegg_paths %>%
  dplyr::group_by(category) %>%
  dplyr::summarise(num = n(), .groups = "drop") %>%
  dplyr::arrange(desc(num)) %>%
  ggplot(aes(x = num, y = fct_reorder(category, num), fill = category)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Category", values = colors_keggcat) +
  labs(title = "KEGG pathways per category", x = "# of pathways", y = "Category name") +
  guides(fill = "none") +
  theme_local() +
  theme(axis.text = element_text(size = 11))
ggsave("enrichKEGG/plots/plot_KEGG_stats_cat.png", height = 5, width = 9)
#ggsave("enrichKEGG/plots/plot_KEGG_stats_cat.pdf", height = 5, width = 7)

# Plot stats of subcategories
kegg_paths %>%
  dplyr::group_by(category, subcategory) %>%
  dplyr::summarise(num = n(), .groups = "drop") %>%
  dplyr::arrange(desc(num)) %>%
  ggplot(aes(x = num, y = fct_reorder(subcategory, num), fill = category)) +
  geom_col(position = "dodge2") +
  geom_text(aes(label = num), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_fill_manual(name = "Subcategory", values = colors_keggcat) +
  labs(title = "KEGG pathways per subcategory",
       subtitle = paste0("Total of subcategories: ", length(unique(kegg_paths$subcategory))),
       x = "# of pathways", y = "Subcategory name") +
  theme_local()
ggsave("enrichKEGG/plots/plot_KEGG_stats_subcat.png", height = 9, width = 9)
```

## KEGG pathway network

Using KEGGgraph package, we explore and manipulate KEGG pathways in a graph-like approach.

```{r keggmapviz}
# Read KEGG pathway file
map_kg <- parseKGML2Graph("KEGGgraph/data/hsa04148.xml", expandGenes = TRUE)
#map_file <- parseKGML("KEGGgraph/data/hsa04148.xml")

# Subset graph
map_sub <- subGraph(sample(nodes(map_kg), 15), map_kg)

# Nodes with in-degree (incoming edges; regulations received) and out-degree (regulations sent)
node_i <- sapply(inEdges(map_sub), length) > 0
node_o <- sapply(edges(map_sub), length) > 0
nodes_ios <- node_o | node_i

# Translate the KEGG IDs into gene names --- or use KOs?
if(suppressMessages(require(org.Hs.eg.db))) {
  node_io <- translateKEGGID2GeneID(names(nodes_ios))
  node_names <- sapply(mget(node_io, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)
} else {
  node_names <- names(nodes_ios)
}
names(node_names) <- names(nodes_ios)

# Function to assign attributes (color) to specific nodes of a graph
makeAttr <- function(graph, default, valNodeList) {
  tmp <- nodes(graph)
  x <- rep(default, length(tmp))
  names(x) <- tmp
  
  if(!missing(valNodeList)) {
    stopifnot(is.list(valNodeList))
    allnodes <- unlist(valNodeList)
    stopifnot(all(allnodes %in% tmp))
    for(i in seq(valNodeList)) {
      x[valNodeList[[i]]] <- names(valNodeList)[i]
    }
  }
  return(x)
}

# Define node attributes
node_attrs <- list();
node_attrs$fillcolor <- makeAttr(map_sub, "lightgrey", list(lightblue = names(nodes_ios)[nodes_ios]))
node_attrs$label <- node_names

# Plot graph (using Rgraphviz)
plot(map_sub, "neato", nodeAttrs = node_attrs,
     attrs = list(node = list(width = "0.75", shape = "ellipse"),
                  edge = list(arrowsize = "0.7")))

# Merge pathways: many KEGG pathways embed other pathway

# Betweenness is a centrality measure of a node within a graph
map_bcc <- RBGL::brandes.betweenness.centrality(map_kg)
map_bccval <- map_bcc$relative.betweenness.centrality.vertices[1L, ]

# Get top 5 genes by betweenness
top_bcc <- sort(map_bccval, decreasing = TRUE)[1:5]
print(top_bcc)

# Convert KEGG IDs to gene names
top_bccnames <- names(top_bcc)
top_bccid <- translateKEGGID2GeneID(top_bccnames)
top_bccgenes <- sapply(mget(top_bccid, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)
print(top_bccgenes)

# Subset graph by near nodes of top genes
top_in <- sapply(top_bccnames, function(x) inEdges(map_kg)[x])
top_out <- sapply(top_bccnames, function(x) edges(map_kg)[x])
top_subnodes <- unique(unname(c(unlist(top_in), unlist(top_out), top_bccnames)))
top_subg <- subGraph(top_subnodes, map_kg)

# Get gene names of subgraph
top_subnodes <- nodes(top_subg)
top_subids <- translateKEGGID2GeneID(top_subnodes)
top_subgenes <- sapply(mget(top_subids, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)

# Plot subgraph of key genes
top_list <- unname(top_bccgenes)
top_lab <- lapply(top_bccnames, function(x) x)
names(top_lab) <- top_list
n_attrs <- list()
n_attrs$label <- makeAttr(top_subg, "", top_lab)
n_attrs$fillcolor <- makeAttr(top_subg, "lightgrey", list(lightblue = top_bccnames))
n_attrs$width <- makeAttr(top_subg, "", list("0.75" = top_bccnames))
#pdf("subgraph_efferocytosis.pdf", width = 9, height = 6)
plot(top_subg, "twopi", nodeAttrs = n_attrs, attrs = list(graph = list(start = 2)))
#dev.off()

# v2
top_subg2 <- top_subg
nodes(top_subg2) <- top_subgenes
plot(top_subg2, "twopi", attrs = list(node = list(fillcolor = "lightblue", font = 5)))

## check -> https://www.kegg.jp/pathway/hsa04148
```

## Gene expression of inflammation

We process RNA-seq data from the GTEx Portal for the inflammatory genes (inflam85).

```{r genexpr1}
# Extract genes from the input list
genexpr_genes <- genes_file %>%
  distinct(HGNC_symbol) %>%
  pull(HGNC_symbol)
print(length(genexpr_genes))

# Convert HGNC symbols to ENSEMBL IDs:
#GTEx Gencode version: v39 -> https://www.gencodegenes.org/human/release_39.html
#ie. ALOX5 is ENSG00000012779.12
mart_genes <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl", version = "105")
genexpr_ensembl <- getBM(attributes = c("hgnc_symbol", "ensembl_gene_id_version"),
                         filters = "hgnc_symbol", values = genexpr_genes, mart = mart_genes)

# Retrieve expression/tissue data from GTEx
genexpr_query <- get_gene_expression(gencodeIds = genexpr_ensembl %>% pull(ensembl_gene_id_version),
                                     datasetId = "gtex_v10", itemsPerPage = 10000) #force GTEx v10
glimpse(genexpr_query)

# Calculate mean TPM and frecuencies per row (genes)
genexpr_tpm <- genexpr_query %>%
  ##check number of samples per gene within each tissue (4165 pass >= 100)
  #rowwise() %>% dplyr::mutate(n_samples = length(unlist(data)))
  dplyr::mutate(sum_tpm = sapply(data, sum, na.rm = TRUE), #sum all samples per gene
                mean_tpm = sapply(data, mean, na.rm = TRUE)) %>% #collapse samples into a mean value
  ##use median instead mean? control zeroes -- median_tpm = sapply(data, median, na.rm = TRUE)
  dplyr::group_by(geneSymbol) %>%
  dplyr::mutate(freq = mean_tpm / sum(mean_tpm), #normalize means to relative freq {0-1}
                freq_abs = mean_tpm) %>% #absolute freq as a sum of TPMs (scale to 10^6?)
  ungroup() %>%
  dplyr::mutate(tissue = str_replace_all(tissueSiteDetailId, "_", " ")) %>% #remove '_'
  dplyr::mutate(freq_log = -freq * log2(freq)) %>% #calc freq log per gene for entropy
  dplyr::select(tissue, ontologyId, gencodeId, geneSymbol, sum_tpm, mean_tpm, freq, freq_abs, freq_log) #drop columns

# Sanity check
print(genexpr_tpm %>%
  dplyr::group_by(geneSymbol) %>%
  dplyr::summarise(total = sum(freq)) %>%
  head()) #must be 1 in all genes

# Export data to files
#saveRDS(genexpr_query, "enrichKEGG/output_data/GTEx_raw_data_85genes.rds", compress = "bzip2")
#write.table(genexpr_tpm, "enrichKEGG/output_data/GTEx_freq_85genes.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

### Plot expression across genes

Exploratory distribution of gene expression across tissues.

```{r genexprplot1}
# Plot distribution across genes
genexpr_tpm %>%
  ggplot(aes(x = tissue, y = freq, group = geneSymbol, color = geneSymbol)) +
  geom_line(linewidth = 0.25, alpha = 0.85) +
  guides(color = guide_legend(title = "Gene")) +
  labs(title = "Gene expression distribution", x = "Tissue", y = "f(gene)") +
  theme_local() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# Plot all genes facetted
genexpr_tpm %>%
  ggplot(aes(x = tissue, y = freq_abs, color = geneSymbol, group = geneSymbol)) +
  geom_line(linewidth = 0.25, alpha = 0.85) +
  labs(title = "Gene expression distribution", x = "Tissue", y = "f(gene)") +
  guides(color = "none", group = "none") +
  facet_wrap(~geneSymbol) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 2, angle = 45, vjust = 1, hjust = 1),
        strip.background = element_blank())
ggsave("enrichKEGG/plots/gene_expression_per_tissue.png", height = 10, width = 18)
```

### Heatmap across enriched pathways

Explore gene expression distribution across tissues and enriched pathways (54 x 208).

```{r genexprhm1}
# Process the gene list from the enrichment results
genexpr_data <- kegg_enrichment %>%
  separate_longer_delim(cols = geneIDs, delim = ",") %>% #
  dplyr::rename(kegg_id = geneIDs) %>% #rename ID
  left_join(genes_file, by = "kegg_id", relationship = "many-to-many") %>% #handle repeated ID: 721
  dplyr::select(pathway, name, kegg_id, HGNC_symbol, count) #drop columns

# Add pathway data
genexpr_path <- genexpr_tpm %>%
  left_join(genexpr_data, by = join_by(geneSymbol == HGNC_symbol), relationship = "many-to-many") %>%
  dplyr::filter(!is.na(pathway)) #drop NA pathways (from 3 KEGG IDs with no pathways assigned)
print(length(unique(genexpr_path$pathway))) #must be 208

# Plot all genes facetted per pathway
genexpr_path %>%
  #dplyr::filter(count > 16) %>%
  dplyr::filter(name == "Efferocytosis") %>%
  ggplot(aes(x = tissue, y = freq, color = pathway, group = geneSymbol)) +
  geom_line(linewidth = 0.25, alpha = 0.85) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(title = "Gene expression distribution per pathway", x = "Tissue", y = "f(gene)") +
  guides(color = "none", group = "none") +
  facet_wrap(~name) +
  theme_local() +
  theme(axis.text.x = element_text(size = 6, angle = 45, vjust = 1, hjust = 1))
ggsave("enrichKEGG/plots/gene_expression_per_pathway.png", height = 7, width = 12)

# Prepare matrix (tissues x pathways)
genexpr_hm <- genexpr_path %>%
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(freq), .groups = "drop") %>%
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Export matrix to file
#write.table(genexpr_hm, "enrichKEGG/output_data/GTEx_matrix_hm_85genes.tsv", sep = "\t", quote = FALSE, row.names = TRUE)

# Set colors for freq values
freq_colors <- colorRamp2(c(0, 0.05, 0.2, 0.5, 1),#c(0, 0.25, 0.5, 0.75, 1),
                          c("white", "#b9c7de", "#5073ad", "#c5849d", "#793850"))

# Plot heatmap
#png("heatmap_tissues_per_pathways.png", height = 15, width = 12, units = "in", res = 300)
#pdf("heatmap_tissues_per_pathways.pdf", height = 15, width = 12)
Heatmap(as.matrix(genexpr_hm), border = TRUE, col = freq_colors,
        row_title = "Pathways", column_title = "Tissues",
        heatmap_legend_param = list(title = "Mean f(gene)",
                                    title_gp = gpar(fontsize = 10, fontface = "bold")),
                                    #at = c(0, 0.25, 0.5, 0.75, 1),
                                    #labels = c(0, 0.25, 0.5, 0.75, 1)),
        row_names_gp = gpar(fontsize = 5), column_names_gp = gpar(fontsize = 8))
#dev.off()
#draw(plot_hm, merge_legend = TRUE)

# Calculate column-wise variances
#genexpr_hm_var <- apply(genexpr_hm, 2, var)

# Keep the 50% most variable columns (tissues)
#genexpr_hm_idx <- order(genexpr_hm_var, decreasing = TRUE)[1:round(0.5 * ncol(genexpr_hm))]
#genexpr_hm_sel <- genes_hm_v2[, genexpr_hm_idx] #match cols
```

Now, we plot the heatmap of the absolute frequencies, where we observe more pronounced differences compared to the previous frequency heatmap:

```{r genexprhm2}
genexpr_path2 <- genexpr_tpm %>%
  left_join(genexpr_data, by = join_by(geneSymbol == HGNC_symbol), relationship = "many-to-many") %>%
  dplyr::filter(!is.na(pathway)) #drop NA pathways (from 3 KEGG IDs with no pathways assigned)

# Prepare matrix (tissues x pathways)
genexpr_hm2 <- genexpr_path2 %>%
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(freq_abs), .groups = "drop") %>%
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Extract significant pathways
genexpr_hm2_signif <- kegg_enrichment %>%
  dplyr::filter(p.adjust <= 0.05) %>%
  dplyr::select(name)
genexpr_hm2_rowcolors <- rep("black", nrow(genexpr_hm2))
genexpr_hm2_rowcolors[rownames(genexpr_hm2) %in% genexpr_hm2_signif$name] <- "red"

# Set colors for freq_abs values
freq_colors2 <- colorRamp2(c(0, 0.25, 0.5, 0.75, 1),
                           c("white", "#b9c7de", "#5073ad", "#c5849d", "#793850"))

# Plot heatmap v2
png("enrichKEGG/plots/heatmap_tissues_per_pathways_v2.png", height = 15, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_hm2), border = TRUE, #col = freq_colors2,
        row_title = "Pathways", column_title = "Tissues",
        heatmap_legend_param = list(title = "Mean f.abs(gene)",
                                    title_gp = gpar(fontsize = 10, fontface = "bold")),#,
                                    #at = c(0, 0.25, 0.5, 0.75, 1),
                                    #labels = c(0, 0.25, 0.5, 0.75, 1)),
        row_names_gp = gpar(fontsize = 5, col = genexpr_hm2_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()
#while (!is.null(dev.list())) dev.off()
```

Explore the effect of the median.

```{r genexprhm3}
# Prepare matrix (tissues x pathways)
genexpr_hm3 <- genexpr_path2 %>%
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(median_freq = median(freq_abs), .groups = "drop") %>%
  pivot_wider(names_from = tissue, values_from = median_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Plot heatmap v3
png("enrichKEGG/plots/heatmap_tissues_per_pathways_v3.png", height = 15, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_hm3), border = TRUE, col = freq_colors2,
        row_title = "Pathways", column_title = "Tissues",
        heatmap_legend_param = list(title = "Median f.abs(gene)",
                                    title_gp = gpar(fontsize = 10, fontface = "bold"),
                                    at = c(0, 0.25, 0.5, 0.75, 1),
                                    labels = c(0, 0.25, 0.5, 0.75, 1)),
        row_names_gp = gpar(fontsize = 5, col = genexpr_hm2_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()
```

## Gene expression in KEGG

We process RNA-seq data from the GTEx Portal for the full set of genes and pathways in KEGG (universe).

```{r genexpr2}
### -- RUN ONCE: this section -- ###
# Extract all human genes
#genexpr_all_genes <- kegg_mapping %>%
#  distinct(gene_id) %>%
#  dplyr::mutate(gene_id = str_replace(gene_id, "hsa:", "")) %>% #remove prefix
#  pull(gene_id)
#print(length(genexpr_all_genes)) #9421

# Convert ENTREZ/KEGG IDs to ENSEMBL IDs
#genexpr_all_ensembl <- getBM(attributes = c("entrezgene_id", "ensembl_gene_id_version"),
#                             filters = "entrezgene_id", values = genexpr_all_genes, mart = mart_genes)
#genexpr_hsa_genes <- genexpr_all_ensembl %>% pull(ensembl_gene_id_version) #to list

# Split the query in smaller chunks
#genexpr_chunks <- split(genexpr_hsa_genes, ceiling(seq_along(genexpr_hsa_genes) / 500))

# Retrieve expression/tissue data from GTEx -> ~30 min, high mem
#genexpr_query_list <- purrr::map(genexpr_chunks,
#                                 ~ get_gene_expression(gencodeIds = .x, #pass each chunk
#                                                       datasetId = "gtex_v10", itemsPerPage = 100000))

# Merge lists of output dataframes
#genexpr_all_query <- do.call(rbind, genexpr_query_list)
#glimpse(genexpr_all_query)
#rm(genexpr_query_list) #size of 1.4G, so remove it

# Calculate mean TPM and frecuencies per row (genes)
genexpr_all_tpm <- genexpr_all_query %>%
  dplyr::mutate(sum_tpm = sapply(data, sum, na.rm = TRUE), #sum all samples per gene
                mean_tpm = sapply(data, mean, na.rm = TRUE)) %>% #collapse samples into a mean value
  dplyr::group_by(geneSymbol) %>%
  dplyr::mutate(freq = mean_tpm / sum(mean_tpm), #normalize means to relative freq {0-1}
                freq_abs = mean_tpm / max(mean_tpm)) %>% #absolute freq (highest expression = 1)
  ungroup() %>%
  dplyr::mutate(tissue = str_replace_all(tissueSiteDetailId, "_", " ")) %>% #remove '_'
  dplyr::mutate(freq_log = ifelse(freq > 0, -freq * log2(freq), 0)) %>% #calc. entropy: S
  dplyr::select(tissue, ontologyId, gencodeId, geneSymbol, sum_tpm, mean_tpm, freq, freq_abs, freq_log) #drop columns

# Export main objects (do this in the server)
#saveRDS(genexpr_all_ensembl, file = "genexpr_all_ensembl.rds", compress = "bzip2")
#saveRDS(genexpr_all_query, file = "genexpr_all_query.rds", compress = "bzip2")
#saveRDS(genexpr_all_tpm, file = "genexpr_all_tpm.rds", compress = "bzip2")
### -- -- ###
```

### Prepare data

Use previous objects to process gene expression data within the KEGG pathways across different tissues.

```{r genexpr2path}
# Load object with TPM and freqs
genexpr_all_tpm <- readRDS("enrichKEGG/data/genexpr_all_tpm.rds")
glimpse(genexpr_all_tpm)

# Exclude cells cultures (non-tissues)
genexpr_all_tpm <- genexpr_all_tpm %>%
  dplyr::filter(tissue != "Cells EBV-transformed lymphocytes" & tissue != "Cells Cultured fibroblasts")
print(length(unique(genexpr_all_tpm$tissue))) #total tissues: 52

# Also, remove the tissues with less than 100 samples?
#dplyr::filter(purrr::map_int(data, length) >= 100) or dplyr::filter(lengths(data) >= 100)

# Load object with gene IDs associations
genexpr_all_ensembl <- readRDS("enrichKEGG/data/genexpr_all_ensembl.rds")
glimpse(genexpr_all_ensembl)
# Some Entrez ID maps to multiple Ensembl versioned genes
#genexpr_all_ensembl %>% count(entrezgene_id) %>% dplyr::filter(n > 1)

# Use symbols to map genes back to Entrez IDs
#genexpr_all_keggids <- getBM(attributes = c("hgnc_symbol", "entrezgene_id"),
#                             filters = "hgnc_symbol", values = genexpr_all_tpm %>% pull(geneSymbol),
#                             mart = mart_genes)

# Add Entrez IDs using ENSEMBL versioned IDs
genexpr_all_tpm <- genexpr_all_tpm %>%
  left_join(genexpr_all_ensembl, by = join_by(gencodeId == ensembl_gene_id_version),
            multiple = "first") %>% #only the first match from the right table
  dplyr::mutate(entrezgene_id = as.character(entrezgene_id), #cast
                entrezgene_id = paste0("hsa:", entrezgene_id)) #add KEGG prefix

# Note that not all genes have expression data..
print(length(unique(genexpr_all_tpm$entrezgene_id))) #total genes: 8615

# Map genes to KEGG pathways
genexpr_all_path <- genexpr_all_tpm %>%
  left_join(kegg_mapping, by = join_by(entrezgene_id == gene_id), relationship = "many-to-many")
print(genexpr_all_path %>% dplyr::filter(is.na(path_id)) %>% nrow()) #sanity check = 0
print(length(unique(genexpr_all_path$path_id))) #must be 372

# Add KEGG categories for pathways
genexpr_all_path <- genexpr_all_path %>%
  left_join(kegg_paths, by = join_by(path_id == pathway)) %>%
  dplyr::rename(pathway = path_id) #rename
print(genexpr_all_path %>%
        dplyr::filter(is.na(name)) %>%
        distinct(pathway) %>%
        arrange(pathway)) #must be empty

# Export processed table to object
#saveRDS(genexpr_all_path, file = "enrichKEGG/output_data/genexpr_all_path.rds", compress = "bzip2")
```

### Heatmap across pathways

Explore gene expression distribution across tissues and KEGG pathways.

```{r genexpr2hm}
# Prepare matrix (tissues x pathways)
genexpr_all_hm <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(freq_abs), .groups = "drop") %>% #add pseudocount
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Export matrix to file
#write.table(genexpr_all_hm, "enrichKEGG/output_data/GTEx_all_KEGG_pathways.tsv", sep = "\t", quote = FALSE, row.names = TRUE)

# Set colors for freq values
#range(genexpr_all_hm)
hm_all_colors <- colorRamp2(c(0, 0.2, 0.4, 0.6),
                            c("white", "#5073ad", "#c5849d", "#793850"))
#c("white", "#96abcd", "#5073ad", "#e2c1ce", "#c5849d", "#793850"))
#hm_all_colors = colorRamp2(seq(log10(1), log10(6e5), length = 4), rev(hcl.colors(4, palette = "blues")))

# Label significant enriched pathways
genexpr_all_hm_rowcolors <- rep("black", nrow(genexpr_all_hm))
genexpr_all_hm_rowcolors[rownames(genexpr_all_hm) %in% genexpr_hm2_signif$name] <- "red"

# Plot heatmap
#pdf("enrichKEGG/plots/KEGG_tissues_per_pathways.pdf", height = 18, width = 12)
png("enrichKEGG/plots/KEGG_tissues_per_pathways.png", height = 18, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_all_hm), border = TRUE, col = hm_all_colors,
        row_title = paste0("Pathways (n = ", nrow(genexpr_all_hm),")"),
        column_title = paste0("Tissues (n = ", ncol(genexpr_all_hm), ")"),
        heatmap_legend_param = list(title = "Abs.freq",
                                    title_gp = gpar(fontsize = 10, fontface = "bold")),
        row_names_gp = gpar(fontsize = 4, col = genexpr_all_hm_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()

# Prepare matrix with log10 scale (tissues x pathways)
genexpr_all_hm2 <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(log10(freq_abs + 1)), .groups = "drop") %>% #add pseudocount
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Set colors for freq values
hm_all_colors2 <- colorRamp2(c(0, 0.05, 0.1, 0.15, 0.2),
                             c("white", "#96abcd", "#5073ad", "#c5849d", "#793850"))

# Plot log10 heatmap
#pdf("enrichKEGG/plots/KEGG_tissues_per_pathways_log10.pdf", height = 18, width = 12)
png("enrichKEGG/plots/KEGG_tissues_per_pathways_log10.png", height = 18, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_all_hm2), border = TRUE, col = hm_all_colors2,
        row_title = paste0("Pathways (n = ", nrow(genexpr_all_hm2),")"),
        column_title = paste0("Tissues (n = ", ncol(genexpr_all_hm2), ")"),
        heatmap_legend_param = list(title = expression(bold("log"[10]*"(freq)")),
                                    title_gp = gpar(fontsize = 10)),
        row_names_gp = gpar(fontsize = 4, col = genexpr_all_hm_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()
```

Plot some filtered heatmaps.

```{r genexpr2filt}
# Prepare a filtered matrix containing non-disease pathways
genexpr_all_hm3 <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::filter(category != "Human Diseases") %>% #remove diseases
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(log10(freq_abs + 1)), .groups = "drop") %>% #add pseudocount
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Label significant enriched pathways
genexpr_all_hm2_rowcolors <- rep("black", nrow(genexpr_all_hm3))
genexpr_all_hm2_rowcolors[rownames(genexpr_all_hm3) %in% genexpr_hm2_signif$name] <- "red"

# Plot filtered heatmap
#pdf("enrichKEGG/plots/KEGG_tissues_per_pathways_nondis.pdf", height = 15, width = 12)
png("enrichKEGG/plots/KEGG_tissues_per_pathways_nondis.png", height = 15, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_all_hm3), border = TRUE, col = hm_all_colors2,
        row_title = paste0("Pathways (n = ", nrow(genexpr_all_hm3),")"),
        column_title = paste0("Tissues (n = ", ncol(genexpr_all_hm3), ")"),
        heatmap_legend_param = list(title = expression(bold("log"[10]*"(freq)")),
                                    title_gp = gpar(fontsize = 10)),
        row_names_gp = gpar(fontsize = 4.5, col = genexpr_all_hm2_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()

# Prepare filtered matrix containing only disease pathways
genexpr_all_hm4 <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::filter(category == "Human Diseases") %>% #select diseases
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(log10(freq_abs + 1)), .groups = "drop") %>% #add pseudocount
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Label significant enriched pathways
genexpr_all_hm3_rowcolors <- rep("black", nrow(genexpr_all_hm4))
genexpr_all_hm3_rowcolors[rownames(genexpr_all_hm4) %in% genexpr_hm2_signif$name] <- "red"

# Plot filtered heatmap
#pdf("enrichKEGG/plots/KEGG_tissues_per_pathways_dis.pdf", height = 7, width = 12)
png("enrichKEGG/plots/KEGG_tissues_per_pathways_dis.png", height = 7, width = 12, units = "in", res = 300)
Heatmap(as.matrix(genexpr_all_hm4), border = TRUE, col = hm_all_colors2,
        row_title = paste0("Pathways (n = ", nrow(genexpr_all_hm4),")"),
        column_title = paste0("Tissues (n = ", ncol(genexpr_all_hm4), ")"),
        heatmap_legend_param = list(title = expression(bold("log"[10]*"(freq)")),
                                    title_gp = gpar(fontsize = 10)),
        row_names_gp = gpar(fontsize = 4.5, col = genexpr_all_hm3_rowcolors),
        column_names_gp = gpar(fontsize = 8))
dev.off()

# Prepare filtered matrix with significant enriched pathways
genexpr_all_hm5 <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::filter(name %in% genexpr_hm2_signif$name | name == "Coronavirus disease") %>% #match
  dplyr::group_by(tissue, name) %>%
  dplyr::summarise(mean_freq = mean(log10(freq_abs + 1)), .groups = "drop") %>% #add pseudocount
  pivot_wider(names_from = tissue, values_from = mean_freq) %>% #pivot tissue to columns
  data.frame(row.names = 1, check.names = FALSE) #set pathways as row names

# Plot filtered heatmap
pdf("enrichKEGG/plots/KEGG_tissues_per_pathways_signif.pdf", height = 7, width = 12)
Heatmap(as.matrix(genexpr_all_hm5), border = TRUE, col = hm_all_colors2,
        row_title = paste0("Pathways (n = ", nrow(genexpr_all_hm5),")"),
        column_title = paste0("Tissues (n = ", ncol(genexpr_all_hm5), ")"),
        heatmap_legend_param = list(title = expression(bold("log"[10]*"(freq)")),
                                    title_gp = gpar(fontsize = 10)),
        row_names_gp = gpar(fontsize = 6), column_names_gp = gpar(fontsize = 8))
dev.off()
```

### Entropy analysis

We calculate the entropy for each pathway after gene aggregation (per category).

```{r genexpr2gauss}
# Aggregate the previously calculated S values (shannon entropy)
genexpr_all_entropy <- genexpr_all_path %>%
  dplyr::filter(!is.na(freq)) %>% #discard genes without expression (NaN)
  dplyr::group_by(pathway, name) %>% #aggregate genes per pathway
  dplyr::summarise(genes = n() / length(unique(genexpr_all_tpm$tissue)), #genes * 1/T
                   entropy = sum(freq_log) / genes, #sum * 1/N
                   .groups = "drop")
glimpse(genexpr_all_entropy)
print(summary(genexpr_all_entropy$entropy))

# Export dataset to table
#write.table(genexpr_all_entropy %>% dplyr::select(pathway, name, entropy, category, subcategory),
#            "enrichKEGG/output_data/KEGG_pathways_entropy.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

# Extract pathways
genexpr_all_kegg <- genexpr_all_path %>%
  distinct(pathway, category, subcategory)
genexpr_all_entropy <- genexpr_all_entropy %>%
  left_join(genexpr_all_kegg, by = "pathway")

# Gaussian modeling (EM: expectation-maximization)
genexpr_all_entropy_fit <- genexpr_all_entropy %>%
  #dplyr::filter(pathway != "01100") %>% #remove outlier
  dplyr::group_by(category) %>%
  dplyr::summarise(model = list(Mclust(entropy)), .groups = "drop")
#table(sapply(genexpr_all_entropy_fit$model, `[[`, "G"))
#2 categories were best fit by a single Gaussian (G = 1)
#4 categories were best fit by a two-component Gaussian mixture (G = 2)

# Function to map density vectors
mixture_density <- function(mod, x) {
  mu <- as.numeric(mod$parameters$mean)
  pro <- as.numeric(mod$parameters$pro)
  sig <- sqrt(as.numeric(mod$parameters$variance$sigmasq))
  
  # If one variance is shared by all components
  if (length(sig) == 1)
    sig <- rep(sig, length(mu))
  y <- numeric(length(x))
  
  for (k in seq_along(mu)) {
    y <- y + pro[k] * dnorm(x, mean = mu[k], sd = sig[k])
  }
  y
}

# Extract multimodal densities
gaussian_densities <- genexpr_all_entropy_fit %>%
  dplyr::mutate(density = purrr::map(model, function(mod) {
    x <- seq(min(mod$data), max(mod$data), length.out = 300)
    tibble(entropy = x, density = mixture_density(mod, x))
  })) %>%
  dplyr::select(category, density) %>%
  unnest(density)

# Get intercepts
#gaussian_intercepts <- gaussian_densities %>%
#  dplyr::group_by(category) %>%
#  dplyr::summarise(left = min(entropy[density > 0.05]),
#                   right = max(entropy[density > 0.05]), .groups = "drop")

# Extract for each model the gaussian components.
# This for both G=1 and G=2: mean ±2σ (95%) or ±3σ (99.7%)
gaussian_params <- genexpr_all_entropy_fit %>%
    dplyr::mutate(components = purrr::map(model, function(mod) {
      mu <- as.numeric(mod$parameters$mean) #mean = max or x-coordinate of the peak
      sd <- sqrt(as.numeric(mod$parameters$variance$sigmasq))
      
      # If a single variance is shared by all components
      if (length(sd) == 1)
        sd <- rep(sd, length(mu))
      
      tibble(component = seq_along(mu), mean = mu, sd = sd,
             left = mu - 2 * sd, right = mu + 2 * sd)
    })) %>%
  dplyr::select(category, components) %>%
  unnest(components)

# Plot for each category the Probability Density Function (PDF)
genexpr_all_entropy %>%
  ggplot(aes(x = entropy, fill = category)) +
  geom_density(color = "#3f75a2", linewidth = 0.25, alpha = 0.6) +
  geom_line(data = gaussian_densities, aes(entropy, density), color = "red", linewidth = 0.4) +
  # Add peak lines
  geom_vline(data = gaussian_params, aes(xintercept = mean),
             color = "black", linetype = "dashed", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = gaussian_params,
                           aes(x = mean, y = 2.25, label = round(mean, 2)),
                           hjust = -0.2, size = 2) +
  # Add mean ranges
  geom_vline(data = gaussian_params, aes(xintercept = left),
             color = "red", linetype = "dashed", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = gaussian_params,
                           aes(x = left, y = 2.25, label = round(left, 2)),
                           hjust = -0.2, size = 2, max.overlaps = Inf, point.padding = 0.1,
                           min.segment.length = 0.1, segment.size = 0.05) +
  geom_vline(data = gaussian_params, aes(xintercept = right),
             color = "red", linetype = "dashed", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = gaussian_params,
                           aes(x = right, y = 2.25, label = round(right, 2)),
                           hjust = -0.2, size = 2, max.overlaps = Inf, point.padding = 0.1,
                           min.segment.length = 0.1, segment.size = 0.05) +
  scale_fill_brewer("Category", palette = "Pastel2") +
  facet_wrap(~category, axes = "all_x", nrow = 2) +
  labs(title = "Entropy distribution across KEGG pathways", x = "Entropy", y = "Density",
       subtitle = paste0("Total pathways: ", nrow(genexpr_all_entropy))) +
  guides(fill = "none") +
  theme_local()
ggsave("enrichKEGG/plots/KEGG_pathway_entropy.png", device = ragg::agg_png, height = 7, width = 10, dpi = 300)
#ggsave("enrichKEGG/plots/KEGG_pathway_entropy.pdf", height = 7, width = 10)

# Prepare data for Cumulative Distribution Function (CDF)
# Compute kernel density
entropy_density <- density(genexpr_all_entropy$entropy)

# Approximate the CDF
entropy_density_cdf <- cumsum(entropy_density$y)
entropy_density_cdf <- entropy_density_cdf / max(entropy_density_cdf) #normalize to end at 1

entropy_density <- data.frame(x = entropy_density$x, cdf = entropy_density_cdf)

entropy_density %>%
  ggplot(aes(x, cdf)) + #fill = category
  geom_line() #stat_ecdf()

# Extract quantiles for entropy values:
# define the bottom 5% as specialists and top 5% as generalists
quantile(genexpr_all_entropy$entropy, probs = c(0.05, 0.95))

# Print which pathways are below 5%
genexpr_all_entropy %>%
  dplyr::filter(entropy < 3.491985) %>%
  arrange(entropy)

# Match these pathways with those significant enriched
intersect(genexpr_all_entropy %>% dplyr::filter(entropy < 3.491985) %>% pull(name),
          kegg_enrichment %>% dplyr::filter(signif == "*") %>% pull(name))
```

We re-run the Gaussian EM modeling constraining the number of Gaussians (G) for each category.

```{r genexpr2gauss2}
# Define the number of gaussians or G value for each category
gaussian_G_by_category <- c(
  "Cellular Processes" = 2,
  "Environmental Information Processing" = 2,
  "Genetic Information Processing" = 2, #3
  "Human Diseases" = 2,
  "Metabolism" = 2,
  "Organismal Systems" = 2
)

# Gaussian modeling, forcing by G value
genexpr_all_entropy_fit2 <- genexpr_all_entropy %>%
  dplyr::group_by(category) %>%
  dplyr::summarise(model = list(Mclust(entropy, G = gaussian_G_by_category[category])),
                   .groups = "drop")

# Extract multimodal densities
gaussian_densities2 <- genexpr_all_entropy_fit2 %>%
  dplyr::mutate(density = purrr::map(model, function(mod) {
    x <- seq(min(mod$data), max(mod$data), length.out = 300)
    tibble(entropy = x, density = mixture_density(mod, x))
  })) %>%
  dplyr::select(category, density) %>%
  unnest(density)

# Extract for each model the gaussian components
gaussian_params2 <- genexpr_all_entropy_fit2 %>%
    dplyr::mutate(components = purrr::map(model, function(mod) {
      mu <- as.numeric(mod$parameters$mean) #mean = max or x-coordinate of the peak
      sd <- sqrt(as.numeric(mod$parameters$variance$sigmasq))
      
      # If a single variance is shared by all components
      if (length(sd) == 1)
        sd <- rep(sd, length(mu))
      
      tibble(component = seq_along(mu), mean = mu, sd = sd,
             left = mu - 2 * sd, right = mu + 2 * sd)
    })) %>%
  dplyr::select(category, components) %>%
  unnest(components)

# Small wrapper function to use text repel
ggrepel_text_local <- function(...) {
  ggrepel::geom_text_repel(
    ...,
    hjust = -0.2,
    size = 2.5,
    max.overlaps = Inf,
    point.padding = 0.2,
    box.padding = 0.5,
    force = 1,
    force_pull = 0.2,
    min.segment.length = 0.1,
    segment.size = 0.05
  )
}

# Plot for each category the PDF
genexpr_all_entropy %>%
  ggplot(aes(x = entropy, fill = category)) +
  geom_density(color = "#3f75a2", linewidth = 0.25, alpha = 0.6) +
  geom_line(data = gaussian_densities2, aes(entropy, density), color = "red", linewidth = 0.4) +
  # Add peak lines
  geom_vline(data = gaussian_params2, aes(xintercept = mean),
             color = "black", linetype = "dashed", linewidth = 0.2) +
  ggrepel_text_local(data = gaussian_params2,
                     aes(x = mean, y = 2, label = round(mean, 2))) +
  # Add mean ranges
  geom_vline(data = gaussian_params2, aes(xintercept = left),
             color = "red", linetype = "dashed", linewidth = 0.2) +
  ggrepel_text_local(data = gaussian_params2, color = "red",
                     aes(x = left, y = 2.125, label = round(left, 2))) +
  geom_vline(data = gaussian_params2, aes(xintercept = right),
             color = "blue", linetype = "dashed", linewidth = 0.2) +
  ggrepel_text_local(data = gaussian_params2, color = "blue",
                     aes(x = right, y = 2.25, label = round(right, 2))) +
  scale_fill_brewer("Category", palette = "Pastel2") +
  facet_wrap(~category, axes = "all_x", nrow = 2) +
  labs(title = "Entropy distribution across KEGG pathways", x = "Entropy", y = "Density",
       subtitle = paste0("Total pathways: ", nrow(genexpr_all_entropy))) +
  guides(fill = "none") +
  theme_local() +
  theme(strip.background = element_blank())
#ggsave("enrichKEGG/plots/KEGG_pathway_entropy_G2.png", device = ragg::agg_png, height = 7, width = 10, dpi = 300)
ggsave("enrichKEGG/plots/KEGG_pathway_entropy_G2.pdf", height = 7, width = 10)

# Detect generalist/specialist pathways
#genexpr_all_entropy %>%
#  dplyr::group_by(category) %>%
#  dplyr::summarise(q5 = quantile(entropy, probs = 0.05),
#                   q95 = quantile(entropy, probs = 0.95), .groups = "drop")
genexpr_all_filt <- genexpr_all_entropy %>%
  dplyr::group_by(category) %>%
  dplyr::filter(entropy < quantile(entropy, probs = 0.05)) %>%
  dplyr::ungroup()

# Match these pathways with those significant enriched
intersect(genexpr_all_filt %>% pull(name),
          kegg_enrichment %>% dplyr::filter(signif == "*") %>% pull(name))

# Check forced Gs
gip <- genexpr_all_entropy %>%
  dplyr::filter(category == "Genetic Information Processing")
fit_G2 <- Mclust(gip$entropy, G = 2)
fit_G3 <- Mclust(gip$entropy, G = 3)

# Higher BIC is better
c(G2_BIC = fit_G2$bic, G3_BIC = fit_G3$bic)

# Check proportions of the components
fit_G2$parameters$pro
fit_G3$parameters$pro

# Plot entropy for G=2 and G=3
#do those pathways are from a meaningful group?
#are they simply the pathways sitting around the 4.8 peak?
plot(fit_G2, what = "density")
plot(fit_G3, what = "density")

# The 2nd group is explaining only 3 pathways
table(fit_G3$classification)

# Look at which pathways correspond to each component
gip %>%
  dplyr::mutate(component = fit_G3$classification) %>%
  dplyr::filter(component == 2) %>%
  dplyr::arrange(entropy)

# Also look at the posterior probability to each component
gip %>%
  dplyr::mutate(component = fit_G3$classification, uncertainty = fit_G3$uncertainty,
                prob_G1 = fit_G3$z[, 1], prob_G2 = fit_G3$z[, 2], prob_G3 = fit_G3$z[, 3]) %>%
  dplyr::filter(component == 2) %>%
  dplyr::arrange(desc(prob_G3))

# Plot the distribution of the enriched pathways
genexpr_all_entropy %>%
  ggplot(aes(x = entropy)) +
  geom_density(aes(x = entropy, color = "Universe", fill = "Universe"),
               linewidth = 0.25, alpha = 0.4) +
  geom_density(data = genexpr_all_entropy %>%
                 dplyr::filter(name %in% genexpr_hm2_signif$name),
               aes(x = entropy, color = "Enriched", fill = "Enriched"),
               linewidth = 0.25, alpha = 0.6) +
  scale_color_manual("Density", values = c("Universe" = "#3f75a2", "Enriched" = "#ff0000")) +
  scale_fill_manual("Density", values = c("Universe" = "#8bacc7", "Enriched" = "#ff9999")) +
  labs(title = "Entropy comparison",
       subtitle = paste0("Total pathways: ", nrow(genexpr_all_entropy),
                         " | Significant enriched: ", nrow(genexpr_hm2_signif)),
       x = "Entropy", y = "Density") +
  theme_local() +
  theme(axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10))

# Plot the distribution comparing diseases vs non-diseases
genexpr_all_entropy %>%
  dplyr::filter(category != "Human Diseases") %>%
  ggplot(aes(x = entropy)) +
  geom_density(color = "#3f75a2", fill = "#8bacc7", linewidth = 0.25, alpha = 0.4) +
  geom_density(data = genexpr_all_entropy %>% dplyr::filter(category == "Human Diseases"),
               color = "#ff0000", fill = "#ff9999", linewidth = 0.25, alpha = 0.6) +
  labs(title = "Entropy comparison: non-disease (blue) vs disease (red)",
       x = "Entropy", y = "Density") +
  theme_local()

# Take a random sample of 100 pathways in explore distribution
```

## Gene Ontology Biological Process

### Map to GOBP

```{r gobp}
library(AnnotationDbi)
library(org.Hs.eg.db)
library(GO.db)

print(columns(org.Hs.eg.db))
print(keytypes(org.Hs.eg.db))

# Convert genes from Entrez IDs to GO IDs
# Get GO Biological Process annotations
genes_goids <- AnnotationDbi::select(org.Hs.eg.db, keys = kegg_enrichment %>%
                                       separate_rows(geneIDs, sep = ",") %>%
                                       distinct(geneIDs) %>% pull(geneIDs),
                                     columns = c("GO", "ONTOLOGY"), keytype = "ENTREZID")

# Keep biological processes
genes_goids <- genes_goids %>% dplyr::filter(ONTOLOGY == "BP")

# Filter out evidence:
# experimental evidence: EXP, IDA, IMP, IGI, IPI

# Explore GO:BP
genes_gobps <- AnnotationDbi::select(GO.db, keys = genes_goids %>% distinct(GO) %>% pull(GO),
                                     columns = c("TERM", "ONTOLOGY"), keytype = "GOID")
#genes_gobps <- genes_gobps %>% left_join(genes_goids)
 
# biomaRt
mart <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")

bp <- getBM(attributes = c("hgnc_symbol", "go_id", "name_1006", "namespace_1003"),
            filters = "hgnc_symbol", values = "ALOX5", mart = ensembl)

bp <- subset(bp, namespace_1003 == "biological_process")

# 1,356 GO terms → semantic/functional clustering → ~10–30 modules → gene × module network
# build a matrix genes x GOs, then use the GO hierarchy to reduce redundancy (GOSemSim::)
```
