---
title: "Catalog_inflammation"
author: "Moises A. Rojas"
date: "08/07/2026"
output:
  html_document:
    toc: true
---

```{r setup}
suppressMessages(library(tidyverse)) #tidy ecosystem
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(ggplot2)) #plotting
suppressMessages(library(patchwork)) #side-by-side plotting
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorspace)) #color gradients
suppressMessages(library(biomaRt)) #ensembl convertion
suppressMessages(library(UniprotR)) #UniprotKB interface
suppressMessages(library(clusterProfiler)) #KEGG enrichment
suppressMessages(library(enrichplot)) #enrichment plots
suppressMessages(library(KEGGREST)) #KEGG id conversion
suppressMessages(library(KEGGgraph)) #KEGG network
suppressMessages(library(Rgraphviz)) #graph subset and merge
```

## A set of inflammatory genes

Convert the initial set of genes (HGNC IDs, n = 85).

```{r convert}
# Connect to the human Ensembl database
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Read file of selected genes as input
genes_file <- read.csv("inflammation_genes.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Create to a list of genes
genes_list <- genes_file %>% pull(HGNC_symbol)

# Map query genes to get IDs from Uniprot and GeneCards
genes_query <- getBM(attributes = c("hgnc_symbol", "uniprotswissprot", "genecards"), #reactome, go_id..
                     filter = "hgnc_symbol", values = genes_list, mart = ensembl)

# Filter empty rows and print results
genes_query <- genes_query %>% filter(uniprotswissprot != "")
print(genes_query)

# Check if some genes are missing..
print(genes_file[which(!(genes_file$HGNC_symbol %in% genes_query$hgnc_symbol)), ])

# Save results to file
genes_query <- full_join(genes_file, genes_query, by = join_by(HGNC_symbol == hgnc_symbol))
#write.table(genes_query, "output.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

First, we need to curate the list. This will allow us to update our internal dataset by updating Sheet nº2 in the following file: https://docs.google.com/spreadsheets/d/1DkwHyp2xs9bxtTHlldugihmQJX4yX36Zi3WXIjvqDoM/edit?gid=1315050263#gid=1315050263.

## Retrieve info from UniprotKB

```{r uniprot}
# UniprotKB
#https://csdaw.github.io/uniprotREST/

# Define the MEFV UniProt Accession ID
#mefv_id <- c("O15553")

# 1. Get General Protein Function (GeneCards equivalent description)
#mefv_function <- GetProteinFunction(ProteinAccList = mefv_id)

# 2. Get Biological Processes and Pathways
#mefv_pathways <- GetPathways_FunctionalAnnotation(ProteinAccList = mefv_id)

# 3. Get Associated Diseases (e.g., Familial Mediterranean Fever)
#mefv_diseases <- GetDiseases_Keywords(ProteinAccList = mefv_id)

# GO terms
# 1. Fetch GO terms from UniProt
#mefv_go <- GetExpression(ProteinAccList = mefv_id)

# 2. Automatically plot the GO terms
# This generates charts highlighting its role in inflammation, the inflammasome, and the cytosol
#Plot.Ontology(mefv_go)
```

## KEGG enrichment analysis (clusterProfiler)

```{r keggenrichcp}
# Read previously curated list of inflammatory genes
genes_file <- read.csv("enrichKEGG/data/curated_genes.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
genes_uniprot <- genes_file %>% pull(uniprot)
print(length(genes_uniprot))

# Database version: Last update 2026/07/13
cat(readLines("https://rest.kegg.jp/info/pathway/"))

# Perform Over-Representation Analysis (ORA) using Uniprot IDs
kegg_res <- enrichKEGG(gene = genes_uniprot, organism = "hsa", keyType = "uniprot",
                       pvalueCutoff = 0.05, pAdjustMethod = "BH", qvalueCutoff = 0.05)

# Plot enriched dotplot
#use: label_format = funlab
#x = ~ -log(p.adjust), 
dotplot(kegg_res, showCategory = 20, font.size = 10)

# Plot barplot
#kegg_res %>%
#  mutate(qscore = -log(p.adjust, base = 10)) %>%
#  barplot(x = "qscore", font.size = 10)

# Plot pathways map
kegg_simmat <- pairwise_termsim(kegg_res, method = "JC")
emapplot(kegg_simmat, showCategory = 20)

# Plot tree
#treeplot(kegg_simmat, showCategory = 20, geneClusterPanel = "pie",
#         cex_category = 0.5, offset_tiplab = 10, pie = "Count")

##---Check in output only pathways or phenotypes?

# Results post-processing
# Custom function to convert ratio strings (e.g., "10/100") to numeric decimals
#parse_ratio <- function(ratio_str) {
#  sapply(ratio_str, function(x) {
#    nums <- as.numeric(strsplit(x, "/")[[1]])
#    return(nums[1] / nums[2])
#  })
#}

# Filter for pathways where enrichment is at least 3-fold higher than background
#ego_filtered <- ego %>%
#  mutate(
#    GeneRatio_num = parse_ratio(GeneRatio),
#    BgRatio_num = parse_ratio(BgRatio),
#    FoldChange = GeneRatio_num / BgRatio_num
#  ) %>%
#  filter(FoldChange >= 3.0)

# Plot the high-fold-change pathways
#dotplot(ego_filtered)
```

## KEGG enrichment analysis (manual curation)

```{r keggenrichmc}
# First, convert list of genes from Uniprot IDs to KEGG IDs
kegg_genes <- bitr_kegg(geneID = genes_uniprot, fromType = "uniprot", toType = "kegg", organism = "hsa")

# Check duplicated entries (genes associated with more than one KEGG ID)
print(unique(kegg_genes$uniprot[duplicated(kegg_genes$uniprot)]))
#kegg_genes <- kegg_genes %>% distinct(kegg, .keep_all = TRUE)

# Append organism prefix to each entry
genes_map <- kegg_genes %>% pull(kegg)
genes_map <- append(genes_map, list(paste0("hsa:", genes_map)))
genes_map <- unlist(tail(genes_map, 1)) #keep last position

# Convert KEGG IDs to KOs
#https://www.genome.jp/dbget-bin/www_bget?hsa:721
kegg_kos <- keggLink(target = "ko", source = genes_map)

# Check duplicated entries
print(kegg_kos[duplicated(kegg_kos)])

# Create table and export it to file
kegg_id_ko <- tibble(ID = names(kegg_kos), KO = unname(kegg_kos))
kegg_id_ko <- kegg_id_ko %>%
  dplyr::mutate(ID = str_replace(ID, "hsa:", ""),
                KO = str_replace(KO, "ko:", ""))
#write.table(kegg_id_ko, "enrichKEGG/data/KEGG_id2ko.tsv", sep = "\t", quote = FALSE, row.names = FALSE)

# Map pathway names based on KOs
kegg_ko_path <- keggLink("pathway", kegg_kos)
kegg_data <- tibble(KO = names(kegg_ko_path), pathway = unname(kegg_ko_path))

# Which KOs are missing? (not associated with any pathway)
print(length(unique(kegg_data$KO))) #81
print(kegg_kos[!kegg_kos %in% kegg_data$KO])

# Modify and filter columns
kegg_data <- kegg_data %>%
  dplyr::mutate(KO = str_replace(KO, "ko:", "")) %>%
  dplyr::mutate(pathway = str_replace(pathway, "path:map", "")) %>% #keep map?
  dplyr::filter(!grepl("ko", pathway))

# Check the pathway names from the KEGG database v. 042022
pathway_list <- read.csv("enrichKEGG/data/pathway.list.mod.tsv", sep = "\t", header = TRUE,
                         colClasses = c("character", "character"), stringsAsFactors = FALSE)
print(kegg_category[!(kegg_category$id %in% pathway_list$id), ]) #ie. 04148 --> Efferocytosis

# Get KEGG pathway names from API to add them
kegg_categories <- clusterProfiler:::kegg_category_data()
kegg_data <- left_join(kegg_data, kegg_categories, by = join_by(pathway == id))

# Edit pathway name
kegg_data <- kegg_data %>%
  dplyr::mutate(name = str_replace(name, "Efferocytosis New!", "Efferocytosis"))

# Process NAs -> https://www.kegg.jp/pathway/map<pathway>
#kegg_data %>% filter(is.na(name)) %>% distinct(pathway)
kegg_data <- kegg_data %>%
  dplyr::mutate(name = case_when(pathway == "04081" ~ "Hormone signaling",
                                 pathway == "04082" ~ "Neuroactive ligand signaling",
                                 pathway == "04517" ~ "IgSF CAM signaling",
                                 pathway == "04518" ~ "Integrin signaling",
                                 pathway == "04519" ~ "Cadherin signaling",
                                 TRUE ~ name)) #subcategory, category

# Get KO name? *** ej. K03989 -> C4; complement component 4

# Plot pathways distribution
kegg_data %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(count_p = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(count_p)) %>%
  dplyr::slice(1:20) %>% #top 20
  ggplot(aes(x = count_p, y = fct_reorder(name, count_p))) + #, fill = name
  geom_col(position = "dodge2", fill = "#459682") +
  #scale_fill_discrete_diverging(name = "", palette = "Purple-Green") +
  geom_text(aes(label = count_p), vjust = 0.5, hjust = -0.25, size = 3) +
  labs(title = "KEGG pathways distribution (top 20)", x = "# of pathways", y = "Pathway name") +
  theme_classic() +
  theme(legend.position = "none")

# Calculate the enrichment-like ratio of genes per pathway
kegg_enrichment <- kegg_data %>%
  dplyr::group_by(KO, pathway, name) %>%
  dplyr::summarise(n_paths = n(), .groups = "drop") %>% #separate by KOs
  ungroup() %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(across(KO, \(x) paste(na.omit(x), collapse = ",")),
                   count = n(), .groups = "drop") %>% #collapse KOs by pathway and count them
  ungroup() %>%
  dplyr::mutate(ratio = count / length(unique(kegg_data$KO))) #gene ratio is count / total genes

# Define a hypergeometric distribution to calculate the p-value
# overlap between the gene set and the dataset
#pvalue <- phyper(k-1, M, N-M, n, lower.tail = FALSE)
# Use Fisher's exact test with correction methods, such as Bonferroni or false discovery rate (FDR)
# It compares the observed distribution to the expected distribution under the null hypothesis of no enrichment
# So, calculate manually the BgRatio (M and N)

# Download the complete KEGG pathway-to-gene mapping
kegg_all_hsa <- keggLink("pathway", "hsa")
kegg_mapping <- data.frame(gene_id = gsub("", "", names(kegg_all_hsa)), #hsa:
                           path_id = gsub("path:hsa", "", unname(kegg_all_hsa)))

# Define the actual background universe
kegg_bg_genes <- unique(kegg_mapping$gene_id)

# Filter KEGG annotations to only include genes present in our universe
kegg_map_filt <- kegg_mapping %>% dplyr::filter(gene_id %in% genes_map)

# Calculate N (total unique annotated genes in our universe)
bg_ratio_N <- length(unique(kegg_map_filt$gene_id))

# Calculate M (per pathway)
bg_ratios <- kegg_map_filt %>%
  dplyr::group_by(path_id) %>%
  dplyr::summarise(M = n_distinct(gene_id), N = bg_ratio_N, BgRatio = paste0(M, "/", N))

# Dotplot of the local enrichment
kegg_enrichment %>%
  dplyr::arrange(desc(ratio)) %>%
  dplyr::slice(1:20) %>% #top 20
  ggplot(aes(x = ratio, y = fct_reorder(name, ratio))) +
  geom_point(aes(size = ratio), color = "#7b4b8f", alpha = 0.85) + #color = p.adjust
  scale_size(range = c(0, 10)) +
  scale_x_continuous(limits = c(0.1, 0.3)) +
  scale_y_discrete(expand = c(0.05, 0)) + #add y-axis space
  #scale_color_gradient(low = "red", high = "blue", name = "p.adjust") +
  #scale_size_continuous(name = "Gene Count") +
  guides(size = "none") +
  labs(title = "KEGG local enrichment (top 20)", x = "Gene (KO) ratio", y = "Pathway name") +
  theme_bw()
# add category, subcategory for facetting?

# Plot top KOs per pathway (PCA?)
```

## KEGG pathway visualization

```{r keggmapviz}
# Read KEGG pathway file
map_kg <- parseKGML2Graph("KEGGgraph/data/hsa04148.xml", expandGenes = TRUE)
#map_file <- parseKGML("KEGGgraph/data/hsa04148.xml")

# Subset graph
map_sub <- subGraph(sample(nodes(map_kg), 15), map_kg)

# Nodes with in-degree (incoming edges; regulations received) and out-degree (regulations sent)
node_i <- sapply(inEdges(map_sub), length) > 0
node_o <- sapply(edges(map_sub), length) > 0
nodes_ios <- node_o | node_i

# Translate the KEGG IDs into gene names --- or use KOs?
if(suppressMessages(require(org.Hs.eg.db))) {
  node_io <- translateKEGGID2GeneID(names(nodes_ios))
  node_names <- sapply(mget(node_io, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)
} else {
  node_names <- names(nodes_ios)
}
names(node_names) <- names(nodes_ios)

# Function to assign attributes (color) to specific nodes of a graph
makeAttr <- function(graph, default, valNodeList) {
  tmp <- nodes(graph)
  x <- rep(default, length(tmp))
  names(x) <- tmp
  
  if(!missing(valNodeList)) {
    stopifnot(is.list(valNodeList))
    allnodes <- unlist(valNodeList)
    stopifnot(all(allnodes %in% tmp))
    for(i in seq(valNodeList)) {
      x[valNodeList[[i]]] <- names(valNodeList)[i]
    }
  }
  return(x)
}

# Define node attributes
node_attrs <- list();
node_attrs$fillcolor <- makeAttr(map_sub, "lightgrey", list(lightblue = names(nodes_ios)[nodes_ios]))
node_attrs$label <- node_names

# Plot graph (using Rgraphviz)
plot(map_sub, "neato", nodeAttrs = node_attrs,
     attrs = list(node = list(width = "0.75", shape = "ellipse"),
                  edge = list(arrowsize = "0.7")))

# Merge pathways: many KEGG pathways embed other pathway

# Betweenness is a centrality measure of a node within a graph
map_bcc <- RBGL::brandes.betweenness.centrality(map_kg)
map_bccval <- map_bcc$relative.betweenness.centrality.vertices[1L, ]

# Get top 5 genes by betweenness
top_bcc <- sort(map_bccval, decreasing = TRUE)[1:5]
print(top_bcc)

# Convert KEGG IDs to gene names
top_bccnames <- names(top_bcc)
top_bccid <- translateKEGGID2GeneID(top_bccnames)
top_bccgenes <- sapply(mget(top_bccid, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)
print(top_bccgenes)

# Subset graph by near nodes of top genes
top_in <- sapply(top_bccnames, function(x) inEdges(map_kg)[x])
top_out <- sapply(top_bccnames, function(x) edges(map_kg)[x])
top_subnodes <- unique(unname(c(unlist(top_in), unlist(top_out), top_bccnames)))
top_subg <- subGraph(top_subnodes, map_kg)

# Get gene names of subgraph
top_subnodes <- nodes(top_subg)
top_subids <- translateKEGGID2GeneID(top_subnodes)
top_subgenes <- sapply(mget(top_subids, org.Hs.egSYMBOL, ifnotfound = NA), "[[", 1)

# Plot subgraph of key genes
top_list <- unname(top_bccgenes)
top_lab <- lapply(top_bccnames, function(x) x)
names(top_lab) <- top_list
n_attrs <- list()
n_attrs$label <- makeAttr(top_subg, "", top_lab)
n_attrs$fillcolor <- makeAttr(top_subg, "lightgrey", list(lightblue = top_bccnames))
n_attrs$width <- makeAttr(top_subg, "", list("0.75" = top_bccnames))
#pdf("subgraph_efferocytosis.pdf", width = 9, height = 6)
plot(top_subg, "twopi", nodeAttrs = n_attrs, attrs = list(graph = list(start = 2)))
#dev.off()

# v2
top_subg2 <- top_subg
nodes(top_subg2) <- top_subgenes
plot(top_subg2, "neato", attrs = list(node = list(fillcolor = "lightblue", font = 5)))

```

## Reactome pathway analysis

```{r reactome}
library(ReactomePA)
# Reactome Enrichment: gsePathway()

```
