---
title: "GWAS_analysis"
author: "Moises A. Rojas"
date: "27/07/2026"
output:
  html_document:
    toc: true
---

```{r setup}
suppressMessages(library(tidyverse)) #tidy ecosystem
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(ggplot2)) #plotting
suppressMessages(library(patchwork)) #side-by-side plotting
suppressMessages(library(RColorBrewer)) #color palettes
suppressMessages(library(pals)) #more colors
suppressWarnings(library(colorspace)) #color gradients
suppressMessages(library(biomaRt)) #ensembl convertion
suppressMessages(library(GenomicRanges)) #genomic positions
suppressMessages(library(arrow)) #read parquet files
suppressMessages(library(ontologyIndex)) #MONDO onthology
suppressMessages(library(disgenet2r)) #disease mapping
```

## GWAS data processing

UK Biobank recruited participants: ~500,000 individuals
Samples used after further QC: 487,442 individuals
  Sample quality
  Population structure
  Relatedness filtering
British ancestry subset: 409,728 individuals
Data sources:
  Questionnaires
  Physical measurements
  Biological samples
  Blood samples
  Genomic data
Process data:
  Filter variants (QC)
  Map variants to genes
    Ensembl VEP
  Map genes to disease onthology
    Open Targets Platform API
  Generate a gene + disease

# Load processed GWAS data

We downloaded raw data from the UK GWAS database (github.com/Nealelab/UK_Biobank_GWAS). Specifically, we selected inflammatory phenotypes (n = 20) and used the 'both_sexes' statistics files. After, quality control (QC) of variants was performed by retaining only significant variants (p-value < 5x10^-8, MAF > 0.01 (1%), and low_confidence_variant = FALSE). Note that the variants are based on Ensembl GRCh37.

```{r read}
gwas_data <- read.csv("GWAS/data/gwas_merged_rs.tsv", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
print(dim(gwas_data))

# Parse phenotypes
gwas_data <- gwas_data %>%
  #filt_20002_1461.gwas.imputed_v3.both_sexes.tsv.gz.tsv -> 20002_1461
  dplyr::mutate(phenotype = str_replace_all(phenotype, "filt_|.gwas.imputed_v3.both_sexes.tsv.gz.tsv", ""))
head(gwas_data)
```

# Map variants to genes

First, we connect to the human Ensembl database of SNPs.

```{r mapgenes}
#mart_genes <- useEnsembl(biomart = "snp", dataset = "hsapiens_snp", version = 116)
mart_snps <- useMart("ENSEMBL_MART_SNP", dataset = "hsapiens_snp")

# Create a list of reference SNPs ID (rsid)
snps_list <- gwas_data %>% pull(rsid)
print(length(snps_list))

snp_mart <- useEnsembl(biomart = "snp", dataset = "hsapiens_snp", host = "https://grch37.ensembl.org")
#https://useast.ensembl.org/index.html

snp_annot <- getBM(attributes = c("refsnp_id", "chr_name", "chrom_start", "allele",
                                  "ensembl_gene_stable_id", "consequence_type_tv"), mart = mart_snps)
write.table(snp_annot, "GRCh37_Ensembl_SNP_annotation.tsv", sep = "\t", quote = FALSE, row.names = FALSE)


# Query genes to get ensembl gene names
snps_query <- getBM(attributes = c("refsnp_id", "phenotype_description", "ensembl_gene_name"),
                    filters = "snp_filter", values = snps_list[1:100], mart = mart_snps) #rs59727397

# Check annotations: many SNP do not map to any gene..
snps_query %>% dplyr::filter(ensembl_gene_name == "")
#snps_ensembl <- snps_query %>% pull(ensembl_gene_id)

# Get gene name and description
mart_genes <- useEnsembl("genes", dataset="hsapiens_gene_ensembl")
snps_query2 <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "description"),
                     filters = "ensembl_gene_id", values = "ENSG00000138892", mart = mart_genes)

# Many GWAS variants are intergenic. For those, we use the nearest gene approach

```

# MONDO

We use the R package ontologyIndex

```{r mondo}
mondo <- get_ontology("GWAS/data/mondo-simple.obo", extract_tags = "minimal")

#mondo$name["MONDO:0004975"]
#"Alzheimer disease" 
#mondo$parents["MONDO:0004975"]
#$`MONDO:0004975`

# Obtain associated diseases (ie. query DisGeNET)


# Map disease IDs to MONDO onthology

#disgenet2
#Sys.setenv(DISGENET_API_KEY = "fbf5160e-978e-4f53-8397-08894abc3f05")
#results <- gene2evidence(gene = "ENSG00000138892", vocabulary = "ENSEMBL", database = "ALL")
#gene2disease(gene = "NLRP3", vocabulary = "MONDO")

```
