---
title: "Inflammatome"
author: "Moises A. Rojas"
date: "31/07/2026"
output:
  html_document:
    toc: true
---

```{r setup}
suppressMessages(library(tidyverse)) #tidy ecosystem
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(ggplot2)) #plotting
suppressMessages(library(patchwork)) #side-by-side plotting
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorRamp2)) #color gradients
suppressMessages(library(biomaRt)) #ensembl convertion
suppressMessages(library(clusterProfiler)) #KEGG enrichment
suppressMessages(library(KEGGREST)) #KEGG id conversion
```

## Load inflammatome dataset

Convert the initial set of genes using the ENTREZ IDs (n = 2,156).

```{r convert}
# Connect to the human Ensembl database
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

# Read input file
inflammatome_file <- read.csv("inflammatome-dataset/data/Inflammatome_genesets_Entrez_2000.tsv",
                              sep = "\t", header = TRUE, stringsAsFactors = FALSE)
glimpse(inflammatome_file)

# Filter top 100 genes separately
inflammatome_file_top <- inflammatome_file %>% dplyr::filter(grepl("top100", gs_name))

# Create a list of genes
inflammatome_list <- inflammatome_file %>% pull(gene)
inflammatome_list_top <- inflammatome_file_top %>% pull(gene)

# Convert genes to ENTREZ ID to KEGG IDs
inflammatome_keggid_top <- bitr_kegg(geneID = inflammatome_list_top, fromType = "ncbi-geneid",
                                     toType = "kegg", organism = "hsa")
```

Typically, in humans the ENTREZ IDs are equivalent to KEGG IDs.

## KEGG enrichment

We use the list of genes with KEGG IDs to map them to KEGG pathways.

### Pathway mapping

```{r keggmap}
# Print database version -> Last update 2026/07/31
cat(readLines("https://rest.kegg.jp/info/pathway/")[[3]])

# Add organism prefix to each entry
inflammatome_keggids <- list(paste0("hsa:", inflammatome_list_top))[[1]]
head(inflammatome_keggids)

# Retrieve gene additional info
inflammatome_keggids_annot <- getBM(attributes = c("entrezgene_id", "hgnc_symbol", "description"),
                                    filters = "entrezgene_id", values = inflammatome_list_top,
                                    mart = ensembl)
inflammatome_keggids_annot <- inflammatome_keggids_annot %>%
  dplyr::mutate(entrezgene_id = as.character(entrezgene_id)) %>%
  dplyr::mutate(description = str_remove(description, "\\s\\[.*$")) #parse description

# We map IDs to pathways
inflammatome_id2path <- keggLink("pathway", inflammatome_keggids)
inflammatome_id2path <- tibble(ID = names(inflammatome_id2path), pathway = unname(inflammatome_id2path))

# Are there KEGG IDs not associated with any pathway?
print(inflammatome_keggids[!inflammatome_keggids %in% inflammatome_id2path$ID])
print(length(unique(inflammatome_id2path$ID))) #71
```

We observe that 29 genes genes are not associated with any KEGG pathway. This reduces the set to 71 KEGG IDs.

```{r id2path}
# Modify and filter columns
inflammatome_id2path <- inflammatome_id2path %>%
  dplyr::mutate(ID = str_replace(ID, "hsa:", "")) %>%
  dplyr::mutate(pathway = str_replace(pathway, "path:hsa", ""))

# Get KEGG pathway names from API
kegg_categories <- clusterProfiler:::kegg_category_data()
inflammatome_id2path <- left_join(inflammatome_id2path, kegg_categories, by = join_by(pathway == id))

# Curate NAs manually
#https://www.kegg.jp/pathway/map<pathway>
#https://www.genome.jp/kegg/docs/upd_kegg.html
print(inflammatome_id2path %>% dplyr::filter(is.na(name)) %>% distinct(pathway))
inflammatome_id2path <- inflammatome_id2path %>%
  dplyr::mutate(category = case_when(pathway == "04081" ~ "Environmental Information Processing",
                                     pathway == "04517" ~ "Environmental Information Processing",
                                     pathway == "04518" ~ "Environmental Information Processing",
                                     pathway == "04519" ~ "Environmental Information Processing",
                                     TRUE ~ category)) %>%
  dplyr::mutate(subcategory = case_when(pathway == "04081" ~ "Signaling molecules and interaction",
                                        pathway == "04517" ~ "Signaling molecules and interaction",
                                        pathway == "04518" ~ "Signaling molecules and interaction",
                                        pathway == "04519" ~ "Signaling molecules and interaction",
                                        TRUE ~ subcategory)) %>%
  dplyr::mutate(name = case_when(pathway == "04081" ~ "Hormone signaling",
                                 pathway == "04517" ~ "IgSF CAM signaling",
                                 pathway == "04518" ~ "Integrin signaling",
                                 pathway == "04519" ~ "Cadherin signaling",
                                 TRUE ~ name))

print(length(unique(inflammatome_id2path$pathway))) #135
print(length(unique(inflammatome_id2path$category))) #6
print(length(unique(inflammatome_id2path$subcategory))) #26
```

Considering this approach, the inflammatome top100 dataset is associated with 135 pathways, distributed across 6 categories and 26 subcategories according to the KEGG classification.

### Distribution plots

```{r keggplot1, fig.width = 10, fig.height = 8, out.width = "100%"}
# Set colors for categories
col_cat <- colorRampPalette(brewer.pal(8, "Set2"))(length(unique(inflammatome_id2path$category)))

# Group by category
pi1 <- inflammatome_id2path %>%
  dplyr::group_by(pathway, category, name) %>%
  dplyr::summarise(n = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n)) %>%
  ggplot(aes(x = n, y = fct_reorder(name, n), fill = category)) +
  geom_col(position = "dodge2", width = 0.75) +
  scale_fill_manual(name = "Category", values = col_cat) +
  scale_x_continuous(limits = c(0, 15)) +
  labs(title = "KEGG pathways per category", x = "# of genes", y = "Pathway name") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 5),
        legend.box.spacing = unit(0.5, "pt"), #space between tree and legend
        legend.key.spacing.x = unit(0.1, "pt"), #space between columns
        legend.key.spacing.y = unit(0.1, "pt"), #space between rows
        legend.spacing.y = unit(0.75, "pt"), #space between legend titles
        legend.key.height = unit(3, units = "mm"),
        legend.key.width = unit(4, units = "mm"))

# Set colors for subcategories
col_subcat <- brewer.dark2(length(unique(inflammatome_id2path$subcategory)))

# Group by subcategory
pi2 <- inflammatome_id2path %>%
  dplyr::group_by(pathway, subcategory, name) %>%
  dplyr::summarise(n = table(pathway), .groups = "drop") %>%
  dplyr::arrange(desc(n)) %>%
  ggplot(aes(x = n, y = fct_reorder(name, n), fill = subcategory)) +
  geom_col(position = "dodge2", width = 0.75) +
  scale_fill_manual(name = "Subcategory", values = col_subcat) +
  scale_x_continuous(limits = c(0, 15)) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(title = "KEGG pathways per subcategory", x = "# of genes", y = "Pathway name") +
  theme_classic() +
  theme(axis.text.y = element_text(size = 5),
        legend.box.spacing = unit(0.5, "pt"), #space between tree and legend
        legend.key.spacing.x = unit(0.1, "pt"), #space between columns
        legend.key.spacing.y = unit(0.1, "pt"), #space between rows
        legend.spacing.y = unit(0.75, "pt"), #space between legend titles
        legend.key.height = unit(3, units = "mm"),
        legend.key.width = unit(4, units = "mm"))

# Agglom. plots
(pi1 / pi2) + plot_annotation(tag_levels = list(c("a", "b")))
ggsave("inflammatome-dataset/plots/inflammatome_KEGG_pathways.png", height = 21, width = 9, dpi = 450)
```

The plots above exhibit the global distribution of the resulting KEGG pathways for the inflammatome top100 dataset. Notably, the most represented pathways are IgSF CAM signaling (14 genes), Chemokine signaling pathway (11 genes), Cytokine-cytokine receptor interaction and Tuberculosis (10 genes each).

### Enrichment

We calculate the enrichment-like ratio of genes per pathway.

```{r enrich}
kegg_inflammatome_enrich <- inflammatome_id2path %>%
  dplyr::group_by(ID, pathway, name) %>% #separate by IDs
  dplyr::summarise(n_paths = n(), .groups = "drop") %>%
  ungroup() %>%
  dplyr::group_by(pathway, name) %>%
  dplyr::summarise(across(ID, \(x) paste(na.omit(x), collapse = ",")),
                   count = n(), .groups = "drop") %>% #collapse IDs by pathway and count them
  ungroup() %>%
  dplyr::rename(geneIDs = ID) %>%
  dplyr::mutate(ratio = count / length(unique(inflammatome_id2path$ID))) #gene ratio is count/total genes
```

Download the complete KEGG human pathway-to-gene mapping.

```{r kegguniv}
kegg_all_hsa <- keggLink("pathway", "hsa")
kegg_mapping <- data.frame(gene_id = gsub("", "", names(kegg_all_hsa)), #hsa:
                           path_id = gsub("path:hsa", "", unname(kegg_all_hsa)))
print(length(unique(kegg_mapping$gene_id))) #total genes
rm(kegg_all_hsa)
```

The KEGG human pathway map to a total of 9,421 genes.

We define the actual background universe of pathways and genes per pathway.

```{r bgpath}
# Define the total number of all unique genes in the background distribution
bg_genes <- length(unique(kegg_mapping$gene_id)) #N
kegg_map_path <- kegg_mapping %>%
  dplyr::group_by(path_id) %>%
  dplyr::summarise(M = n()) #number of genes in the background universe (per pathway)

# Merge background data with the enrichment
kegg_inflammatome_enrich <- kegg_inflammatome_enrich %>%
  left_join(kegg_map_path, by = join_by(pathway == path_id)) %>%
  dplyr::mutate(BgRatio = paste0(M, "/", bg_genes)) #calculate the BgRatio = M/N
```

Using the background universe, a hypergeometric distribution is applied to calculate the p-value. This statistical test evaluates the significance of the overlap between the gene set and the universe.

```{r bgratio}
# Define the number of genes in the gene set
n_paths <- length(unique(inflammatome_id2path$pathway)) #n
#phyper(k-1, M, N-M, n, lower.tail = FALSE)
kegg_inflammatome_enrich <- kegg_inflammatome_enrich %>%
  dplyr::mutate(pvalue = phyper(count - 1, M, bg_genes - M, n_paths, lower.tail = FALSE)) %>%
  dplyr::mutate(p.adjust = p.adjust(pvalue, method = "BH")) %>%
  dplyr::mutate(signif = ifelse(p.adjust < 0.05, "*", "")) #mark significant pathways

# Get and sort the significant pathways
print(kegg_inflammatome_enrich %>%
        dplyr::filter(p.adjust <= 0.05) %>%
        dplyr::select(name, ratio, p.adjust) %>%
        dplyr::arrange(desc(ratio)))
```

We create a dotplot of the local enrichment.

```{r plotenrich}
kegg_inflammatome_enrich %>%
  dplyr::arrange(desc(ratio)) %>%
  dplyr::slice(1:50) %>%
  ggplot(aes(x = ratio, y = fct_reorder(name, ratio))) +
  geom_point(aes(size = count, color = p.adjust), alpha = 0.75) +
  geom_text(aes(label = signif, x = -0.005),size = 4, color = "#3c4d76") + #p.adj
  scale_y_discrete(expand = c(0.05, 0)) + #add y-axis space
  scale_color_gradient(name = "p.adjust", low = "red", high = "blue") +
  guides(size = guide_legend(title = "# genes", order = 1, reverse = TRUE,
                             override.aes = list(color = "black", alpha = 0.85))) +
  labs(title = "KEGG enrichment (top 50)", x = "Gene ratio", y = "Pathway name") +
  theme_classic() +
  theme(legend.key.height = unit(3, units = "mm"),
        legend.key.width = unit(4, units = "mm"))
ggsave("inflammatome-dataset/plots/KEGG_enrichment_top100.png", height = 7, width = 7)
```

Of the top 50 most enriched pathways, only 6 showed statistically significant adjusted p-values (indicated with an asterisk *). Among the top five pathways, "Cytokine-cytokine receptor interaction" displayed a high enrichment ratio despite having lower representation in the background universe (p = 0.141).

```{r enrichgenes}
kegg_inflammatome_genes <- kegg_inflammatome_enrich %>%
  separate_longer_delim(cols = geneIDs, delim = ",") %>% #split 114548,1535,1536,29108,..., to rows
  group_by(geneIDs) %>%
  summarise(genes = n()) %>%
  left_join(inflammatome_id2path, by = join_by(geneIDs == ID)) #add KEGG data

# Plot all genes across pathways
kegg_inflammatome_genes %>%
  distinct(geneIDs, .keep_all = TRUE) %>% #collapse list
  left_join(inflammatome_keggids_annot, by = join_by(geneIDs == entrezgene_id)) %>% #add gene info
  dplyr::mutate(label = paste0(hgnc_symbol, " (KEGG ID: ", geneIDs, ") - ", description)) %>% #label
  arrange(desc(genes)) %>%
  ggplot(aes(x = genes, y = fct_reorder(label, genes))) +
  geom_point(aes(size = genes, color = category), alpha = 0.75) +
  #geom_text(aes(label = genes), vjust = 0.5, hjust = -0.25, size = 3) +
  scale_y_discrete(expand = c(0.025, 0)) + #add y-axis space
  scale_fill_manual(name = "Category", values = colors_cat) +
  guides(size = "none") +
  labs(title = "Inflammatome genes", x = "# of pathways", y = "Gene name and function") +
  theme_classic() +
  theme(legend.box.spacing = unit(0.5, "pt"), #space between tree and legend
        legend.key.spacing.x = unit(0.1, "pt"), #space between columns
        legend.key.spacing.y = unit(0.1, "pt"), #space between rows
        legend.spacing.y = unit(0.75, "pt"), #space between legend titles
        legend.key.height = unit(3, units = "mm"),
        legend.key.width = unit(4, units = "mm"))
ggsave("inflammatome-dataset/plots/KEGG_key_genes_top100.png", height = 15, width = 12)
```

The plot above shows the global distribution of the 71 inflammatome top100 genes and summarizes the pathways identified by the enrichment analysis.

## Overlap to 85 gene set

```{r overlap}

```
