---
title: "RNA-seq"
author: "Moises A. Rojas"
date: "18/08/2026"
output:
  html_document:
    toc: true
---

```{r setup}
suppressMessages(library(tidyverse)) #tidy ecosystem
suppressMessages(library(ggplot2)) #plotting
suppressMessages(library(patchwork)) #side-by-side plotting
suppressMessages(library(RColorBrewer)) #color palettes
suppressWarnings(library(colorRamp2)) #color gradients
suppressMessages(library(biomaRt)) #ensembl convertion
suppressMessages(library(remart)) #ensembl convertion v2
suppressWarnings(library(ggpubr)) #stats & plots
suppressWarnings(library(ggbeeswarm)) #scatter plots
suppressPackageStartupMessages(library(ComplexHeatmap)) #heatmap plot
suppressMessages(library(edgeR)) #DGE analysis

# ggplot(...) + ... + theme_local()
theme_local <- function() {
  theme_classic() +
    theme(
      legend.box.spacing = unit(0.5, "pt"), #space between plot and legend
      legend.key.spacing.x = unit(0.1, "pt"), #space between columns
      legend.key.spacing.y = unit(0.1, "pt"), #space between rows
      legend.spacing.y = unit(0.75, "pt"), #space between legend titles
      legend.key.height = unit(3, units = "mm"),
      legend.key.width = unit(4, units = "mm")
    )
}
```

## Import DGE data

```{r import}
# Read edgeR object (generated on the cluster)
dge <- readRDS("RNA-seq/data/neurodeg_DGE.rds")
glimpse(dge)
```

## Data post-processing

### Load metadata

```{r sramd}
# Load metadata obtained from the NCBI SRA (python codes)
metadata <- read.csv("RNA-seq/data/metadata_harmonized.csv", sep = ",", header = TRUE)

# Extract sample/experiment IDs in DGE
srxs <- rownames(dge$samples)

# Combine data
srxs <- tibble(experiment = srxs)
srxs <- left_join(srxs, metadata, by = "experiment")

# There are repeated entries?
#srxs %>% add_count(experiment) %>% filter(n > 1)
#srxs <- srxs %>% distinct(experiment, .keep_all = TRUE)

# Add groups as treatments
srxs <- left_join(srxs, dge$samples %>%
                    rownames_to_column("experiment") %>%
                    dplyr::select(experiment, group),
                  by = "experiment") %>%
  dplyr::select(where(~ !all(is.na(.) | . == ""))) #remove columns with NA values

# Explore metadata:
srxs %>%
  filter(age != "" & age != "--") %>%
  dplyr::select(SRAStudy, experiment, library_layout, library_selection, tissue, age, sex, group) %>%
  arrange(SRAStudy) %>%
  knitr::kable("simple") 
```

### Fill metadata

```{r fillmd}
# Load cohort -filtered- data
design <- read.csv("RNA-seq/data/neurodeg_design.tsv", sep = "\t", header = TRUE)
design <- design %>% rownames_to_column("experiment")
glimpse(design)

# Save sample IDs without `Tissue_subtype`
samples_notsub <- design %>%
  dplyr::filter(is.na(Tissue_subtype)) %>%
  dplyr::select(experiment) %>%
  pull()

# Function to parse `tissue` from the extracted metadata
fx_parse_tissue <- function(x) {
  case_when(
    # set keywords to return the string after '|'
    str_detect(x, regex("control|disease|,", ignore_case = TRUE)) &
      str_detect(x, fixed("|")) ~ str_trim(str_extract(x, "(?<=\\|).*")),
    # everything else with '|' return the string before the pattern
    str_detect(x, fixed("|")) ~ str_trim(str_extract(x, "^[^|]+")),
    TRUE ~ x #return the original string instead
  )
}

# Populate `tissue_subtype` column
srxs_tissuesub <- srxs %>%
  dplyr::filter(experiment %in% samples_notsub) %>% #only NA subtypes
  dplyr::select(SRAStudy, experiment, tissue) %>%
  dplyr::mutate(tissue_subtype = fx_parse_tissue(tissue)) #parse tissue
# Force first letter to capital
srxs_tissuesub <- srxs_tissuesub %>%
  dplyr::mutate(tissue_subtype = str_replace(tissue_subtype, "^([a-z])", ~ str_to_upper(.x)))

# Add parsed metadata
design <- design %>%
  left_join(srxs_tissuesub %>% dplyr::select(experiment, tissue_subtype), by = "experiment") %>%
  #use `Tissue_subtype` if it is not NA; otherwise use `tissue_subtype`
  dplyr::mutate(Tissue_subtype = coalesce(Tissue_subtype, tissue_subtype)) %>%
  dplyr::select(-tissue_subtype)

# Sanity check (wiki tables too)
#table(design$Tissue_subtype, useNA = "ifany")
design %>%
  count(Tissue, Tissue_subtype) %>%
  knitr::kable("simple")
design %>%
  count(sra_study_acc, Tissue, Tissue_subtype, Disease_subtype) %>%
  arrange(Tissue, sra_study_acc)
```

## Differential gene expression

### Sample-level QC

First, we filter the samples by library size.

```{r samqc}
# Explore the lib size distribution
summary(dge$samples$lib.size)
dge$samples[dge$samples$lib.size < 1e5, ]

# Remove extremely low library sizes (min 100,000 reads)
keep_libsize <- dge$samples$lib.size >= 1e5
dge <- dge[, keep_libsize]
```

### Normalize library size

Then, we normalize the library sizes of the samples.

```{r norm}
# Filter low-expression genes across samples
keep_low <- filterByExpr(dge, group = dge$samples$group)
table(keep_low)
dge <- dge[keep_low, , keep.lib.sizes = FALSE]

# Normalize lib size by the trimmed mean of M-values
dge <- calcNormFactors(dge, method = "TMM")
head(getNormLibSizes(dge))

# Check outlier samples again
dge$samples %>%
  arrange(desc(norm.factors)) %>%
  head()
outlier_samples <- dge$samples %>%
  rownames_to_column("experiment") %>%
  dplyr::filter(norm.factors > 5) %>%
  dplyr::select(experiment) %>%
  pull()

if (length(outlier_samples) > 0) {
  # Remove outlier samples
  dge <- dge[, !colnames(dge) %in% outlier_samples]
  dge$samples %>%
    arrange(desc(norm.factors)) %>%
    head()
  # Iterate
  print("Iterating normalization...")
  keep_low <- filterByExpr(dge, group = dge$samples$group)
  table(keep_low)
  dge <- dge[keep_low, , keep.lib.sizes = FALSE]
  #need to norm again:: dge <- calcNormFactors(dge, method = "TMM")
  #dge$samples %>%
  #  arrange(desc(norm.factors)) %>%
  #  head()
  print(" done.")
  outlier_samples <- vector()
} else {
  print("Iteration performed.")
}
```

Explore the distribution of library sizes in the filtered samples.

```{r plotdist}
# Histogram plot
dge$samples %>%
  ggplot(aes(x = group, y = lib.size, color = group)) +
  geom_quasirandom(size = 3, alpha = 0.8) +
  geom_boxplot(fill = NA, outlier.size = 0, outliers = FALSE) +
  scale_y_log10(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  scale_color_manual(values = c("#98d277", "#b15928")) +
  stat_compare_means(method = "t.test", label = "p.signif", ref.group = ".all.",
                     vjust = 0.1, bracket.size = 0.25, tip.length = 0.01, size = 3,
                     comparisons = list(c("Healthy", "Neurodegenerative"))) +
  guides(color = "none") +
  labs(title = "Library size distribution in the filtered samples",
       subtitle = paste0("Samples that passed filters: ", ncol(dge)),
       x = "Group", y = "Library size (log10)") +
  theme_local()
#ggsave("libsizedist_SRA16.pdf", height = 5, width = 5)
```

### Pipeline use case

Use case filtering by an individual study: Brain - Alzheimer’s disease.

```{r usecase}
# SRA options: SRP034831, or SRP056604 (*keeps all initial number of samples)
usecase <- design %>%
  #dplyr::filter(Tissue == "Brain") %>%
  #dplyr::filter(sra_study_acc == "SRP034831" | sra_study_acc == "SRP056604"
  #              | sra_study_acc == "SRP061397" | sra_study_acc == "SRP142295") %>%
  dplyr::filter(sra_study_acc == "SRP142295") %>%
  dplyr::select(experiment) %>%
  pull()
# Remove sample SRX1114576 belonging to HD from study SRP061397
usecase <- usecase[!usecase %in% "SRX1114576"]

# Filter samples by selected study, then normalize
dge2 <- dge[, colnames(dge) %in% usecase]
keep_low2 <- filterByExpr(dge2, group = dge2$samples$group)
table(keep_low2)
dge2 <- dge2[keep_low2, , keep.lib.sizes = FALSE]
dge2 <- calcNormFactors(dge2, method = "TMM")
table(dge2$samples$group)

# Add variables to the model
dge2$samples$study <- factor("SRP142295")
dge2$samples$tissue <- factor("Brain")
table(dge2$samples$study, dge2$samples$group)
glimpse(dge2$samples)
```

### Normalize counts

Then, we normalize the matrix of raw counts.

```{r cpm}
# Use log2 counts per million
cpms <- cpm(dge2, log = TRUE)
glimpse(cpms)
```

### Perform DGE

Perform differential gene expression (DGE) analysis using the edgeR Quasi-Likelihood pipeline.

```{r dge}
# GLM approach
# 0+: only include a column for each group, not the intercept
# Q1: What is the overall neurodegenerative signature?
# ~ Tissue + sra_study_acc + group
model <- model.matrix(~0 + group, data = dge2$samples)
colnames(model) <- levels(dge2$samples$group)
glimpse(model)

# Make contrasts
contrast <- makeContrasts(Neurodegenerative - Healthy, levels = model)

# Estimate dispersions (as a negative binomial distribution)
dge2 <- estimateDisp(dge2, model)
dge2$common.dispersion
plotBCV(dge2)

#plotMDS(dge2, labels = dge2$samples$group, col = ifelse(dge2$samples$group == "Healthy", "steelblue", "firebrick"))

# Perform quasi-likelihood F-tests
fit <- glmQLFit(dge2, model, robust = TRUE)
plotQLDisp(fit)
res <- glmQLFTest(fit, contrast = contrast)
topTags(res)
```

### Significant genes

* Interpretation of logFC:
- 2^logFC: >1 fold higher in Neurodegenerative than Healthy
- otherwise, if a gene had <0, ie. -0.25 means 4-fold lower in N than H

* Interpretation of F statistic:
- measures how strongly the data support a difference between the two groups
- large F is stronger evidence that the groups differ
- is not directional

* Criteria for DEGs were as follows:
- adjusted p-value < 0.05 and |log2(FC)| > 1

```{r signif}
# Save the FDR values
res_table <- topTags(res, n = Inf, adjust.method = "BH")$table
table(res_table$FDR < 0.05)

# Explore
res_table %>%
  arrange(FDR) %>%
  slice(1:10) %>%
  dplyr::select(gene_id, gene_name, logFC, F, PValue, FDR)

# Get only significant genes
signif <- res_table[res_table$FDR < 0.05 & abs(res_table$logFC) > 1, ]

# Annotate signif genes? ie. NLRP
#topTable(fit, adjust = "BH", resort.by = "P")
```

### Volcano plot

Plot DGEs, where the x-axis is log2|FC| and the y-axis is -log10(p-value). Red dots represent up-regulated genes, blue dots represent down-regulated genes, and black dots represent non-significantly expressed genes.

```{r volcano}
# Mark upreg/downreg genes
res_table$signif <- "Not significant"
res_table$signif[res_table$FDR < 0.05 & res_table$logFC > 1] <- "Upregulated" #p<0.01
res_table$signif[res_table$FDR < 0.05 & res_table$logFC <= -1] <- "Downregulated"
table(res_table$signif)

#res_table$upreg <- res_table$FDR < 0.05 & res_table$logFC >= 1
#res_table$downreg <- res_table$FDR < 0.05 & res_table$logFC <= -1
#res_table$signif <- "Not significant"
#res_table$signif[res_table$upreg] <- "Upregulated"
#res_table$signif[res_table$downreg] <- "Downregulated"
#log2FC (-0.5, 0.5) ?

# Plot with dot labels
#plot_volcano_v1 <- 
res_table %>%
  ggplot(aes(x = logFC, y = -log10(FDR))) +
  geom_point(aes(color = signif), alpha = 0.5, size = 1.5) +
  geom_vline(xintercept = c(-1, 1), color = "grey50", linetype = "dashed", linewidth = 0.2) +
  geom_hline(yintercept = -log10(0.05), color = "grey50", linetype = "dashed", linewidth = 0.2) +
  ggrepel::geom_text_repel(data = res_table %>% dplyr::filter(signif != "Not significant"),
                           aes(x = logFC, y = -log10(FDR), label = gene_name),
                           hjust = -0.2, size = 2, max.overlaps = Inf, point.padding = 0.1,
                           min.segment.length = 0.1, segment.size = 0.05) +
  scale_color_manual("", values = c("Not significant" = "grey70",
                                    "Upregulated" = "red", "Downregulated" = "blue")) +
  labs(title = "Neurodegenerative disease vs Healthy controls",
       subtitle = paste0("Total SRA studies: ", length(unique(design$sra_study_acc)),
                         "\nTotal samples: ", ncol(dge)),
       x = expression("log"[2]*"(fold change)"),
       y = expression("-log"[10]*"(adj. p-value)")) +
  theme_local()
#ggsave("volcano_SRA16.pdf", plot_volcano_v1, height = 7, width = 9)
```

### Gene annotation

Annotate DGE genes.

```{r annot}
# Check pseudogenes, ie PRAMEF31P
mart <- useEnsembl("genes", "hsapiens_gene_ensembl")
genes <- signif$gene_id
annotation <- getBM(
  attributes = c("ensembl_gene_id_version", "external_gene_name", "gene_biotype", "description"),
  filters = "ensembl_gene_id_version",
  values = "ENSG00000223972.5",
  mart = mart
)

# User remart instead of biomaRt
# Remove ensembl version from genes: ENSG00000223972.5 -> ENSG00000223972
remart::getBM(attributes = c("hgnc_symbol", "ensembl_gene_id",
                             "external_gene_name", "transcript_biotype", "gene_biotype",
                             "description"),
              filters = "ensembl_gene_id", values = "ENSG00000223972")


annotation[grepl("pseudogene", annotation$gene_biotype), ]


```

### Heatmap

Heatmap of DGEs: genes (rows) x group (cols) and samples

```{r heatmap}
# Filter the top 20 genes by p-value
top <- head(rownames(res$table)[order(res$table$PValue)], 20)
matrix <- cpm(dge2[top, ], log = TRUE, prior.count = 1)
rownames(matrix) <- dge2$genes$gene_name[match(rownames(matrix), dge2$genes$gene_id)]

# Plot heatmap
pheatmap::pheatmap(matrix, scale = "row", show_colnames = FALSE,
                   annotation_col = dge2$samples[, "group", drop = FALSE])
```

### Multivariate analysis

PCA/UMAP/clustering of significant DGE genes.

```{r pca}

```
